---
title: "generating_hier_data"
author: "Sigrid Bom"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
# 1. Ensure 'pacman' is installed
if (!require("pacman")) install.packages("pacman")

# 2. Load required packages using pacman for efficient package management - ensures packages are installed if missing, then loads them
pacman::p_load(
  tidyverse,    # Core suite for data manipulation (dplyr) and plotting (ggplot2)
  here,         # For creating robust file paths relative to the project root (optional but good practice)
  posterior,    # Provides tools for working with posterior distributions
  cmdstanr,     # The R interface to Stan (via CmdStan)
  bayesplot,    # Specialized plotting functions for Bayesian models
  patchwork,    # For easily combining multiple ggplot plots
  priorsense,   # For automated prior sensitivity analysis
  SBC,          # Simulation-Based Calibration Checks
  future,       # Parallel processing
  furrr,        # Parallel mapping
  pracma,       # to be used in the probit regression formula
  cli,
  devtools,     # to download package from github
  progressr,     # to get progress bar in SBC
  remotes,       # to install specific version of ggplot needed to make priorsense dens plot
  msm
)

devtools::install_github("hyunjimoon/SBC") # to make SBC
#progressr::handlers(global = TRUE) 
#progressr::handlers(c("rstudio", "progress")) # måske bedre??


# priorsense requires specific version of ggplot
#packageVersion("ggplot2")
install.packages("remotes")
remotes::install_version("ggplot2", version = "3.5.1")


# Set a default theme for ggplot for consistency
theme_set(theme_classic()) # Use a classic theme with white background

# Define relative paths 
stan_model_dir <- here("stan")
stan_results_dir <- here("simmodels")
sim_data_dir <- here("simdata")

# Create directories if they don't exist (important for first run)
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir)
if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)
if (!dir.exists(sim_data_dir)) dir.create(sim_data_dir)

# Define path for saving/loading the fitted model object that includes the generated quantities
model_file <- file.path(stan_results_dir, "2_simple_hrd_model_hier_half_normal_half_normal_priors.rds")
sbc_path <- file.path(stan_results_dir, "2_simple_hrd_model_hier_sbc_half_normal_priors.rds")

# Set to TRUE to rerun everything (takes time!), FALSE to load saved results.
regenerate_simulations <- FALSE 
simulate_more_SBC = TRUE

# define functions

# probit regression - psychometric function
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
 }

# inverse logit
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```



# 1. Generating model - same as before but now the priors are arguments in the function

```{r}
simulate_hierarchical_hrd_data <- function(n_agent = NULL,
                                           n_trials = NULL,
                                           prior_mu_alpha_loc = NULL,
                                           prior_mu_alpha_scale = NULL,
                                           prior_sigma_alpha_loc = NULL,
                                           prior_sigma_alpha_scale = NULL,
                                           prior_mu_b_log_loc = NULL,
                                           prior_mu_b_loc_scale = NULL,
                                           prior_sigma_b_log_loc = NULL,
                                           prior_sigma_b_log_scale = NULL,
                                           prior_mu_lambda_logit_loc = NULL,
                                           prior_mu_lambda_logit_scale = NULL,
                                           prior_sigma_lambda_logit_rate = NULL,
                                           seed = NULL
                                           ){
 
  if (!is.null(seed)) set.seed(seed)  # this generates a new seed for every run of the function if the seed is not set. so the only way to get the same results everytime is to set a seed when running the function (like sim_data(..., seed = 1997))
  
  # 1. sample population hyperparameters (if not provided)
  mu_alpha <- rnorm(1, mean = prior_mu_alpha_loc, sd = prior_mu_alpha_scale) # skal outputte omkring -10  # ændret fra 7 til 15 - 

  sigma_alpha <- abs(rnorm(1, mean = prior_sigma_alpha_loc, sd = prior_sigma_alpha_scale))
  # skal outputte omkring 10

  mu_b_log <- rnorm(1, mean = prior_mu_b_log_loc, sd = prior_mu_b_loc_scale) # skal outputte omkring 2.5 
  sigma_b_log <- abs(rnorm(1, mean = prior_sigma_b_log_loc, sd = prior_sigma_b_log_scale)) # skal outputte omrking 0.4 
  
  mu_lambda_logit <- rnorm(1, mean = prior_mu_lambda_logit_loc, sd = prior_mu_lambda_logit_scale) # skal være omrking -3
  sigma_lambda_logit <- rexp(1, rate = prior_sigma_lambda_logit_rate) #skal outputte omkring 0.4
  
  # 1.5 sample individual parameter values using hyper parameter values from population level
  
  data_sim = map_dfr(1:n_agent, ~ {
    # 1. sample true parameters per agent
    alpha         <- rnorm(1, mu_alpha, sigma_alpha)
    b_log         <- rnorm(1, mu_b_log, sigma_b_log)
    beta          <- exp(b_log)
    lambda_logit <- rnorm(1, mu_lambda_logit, sigma_lambda_logit)
    lambda       <- 0.5 * inv_logit(lambda_logit)
  
    # 2. sample stim
    deltaBPM <- msm::rtnorm(n_trials, mean = alpha, sd = 2 * beta, lower = -50, upper = 50) # truncated normal # is it a bad thing to depend the stimuli on the sampled parameters? also should this be different for each participant like it is now?
  
    # 3. calculate theta and choice
    theta_list <- probit_regression(x = deltaBPM, alpha = alpha, beta = beta, lambda = lambda)
    choice = rbinom(n=n_trials, size = 1, prob = theta_list) 
  
    # 4. return a tibble for each agent
    tibble(
      n_agent   = .x,
      n_trials,
      alpha,
      mu_alpha,
      sigma_alpha,
      beta,
      mu_b_log,
      sigma_b_log,
      lambda,
      mu_lambda_logit,
      sigma_lambda_logit,
      deltaBPM,
      theta = theta_list,
      choice
    )
})
  
 return(data_sim) 
}

```


```{r}
hier_data = simulate_hierarchical_hrd_data(n_agent = 60,
                                           n_trials = 40,
                                           prior_mu_alpha_loc = -10,
                                           prior_mu_alpha_scale = 7,
                                           prior_sigma_alpha_loc = 10,
                                           prior_sigma_alpha_scale = 3,
                                           prior_mu_b_log_loc = 2.5,
                                           prior_mu_b_loc_scale = 0.2,
                                           prior_sigma_b_log_loc = 0.4,
                                           prior_sigma_b_log_scale = 0.1,
                                           prior_mu_lambda_logit_loc = -3,
                                           prior_mu_lambda_logit_scale = 0.4,
                                           prior_sigma_lambda_logit_rate = 2.5)

hier_data
```

Justification of prior values:
- We are running the data collection with 60 participants and 40 trials of the HRD. 
- *priors of mu_alpha*: we know from empirical data that participants tend to underestimate their heartrate with a threshold (alpha) of around -10, i.e., why the mean of the prior_mu_alpha is -10. Naturally, this varies, and the SD of 7 allows for variation. 
- *priors of sigma_alpha*: we want between-subject variation to allow for positive thresholds, too, why this is set to 15 (quite broad).

with alpha ~N(mu_alpha, sigma_alpha) and mu_alpha = -10, the probability an agent has an alpha > 0 is:

```{r}
1-pnorm(0, mean = -10, sd = 15)
```

This allows for 25% of agents (i.e., 15 agents out of 60) to have a positive threshold.

- *priors of mu_b_log and sigma_b_log*: 
beta is exp(b_log) and b_log ~ N(mu_b_log, sigma_b_log) per agent, so we want b_log to output around 2.5 +- 0.8 which we get by setting mu_b_log~ N(2.5, 0.2) and sigma_b_log ~N(0.4,1), so beta = exp(b_log) ranges roughly 5.5 to 27, all giving genuine sigmoid shapes over a ±50 stimulus range instead of degenerate step/flat functions.


```{r}


hier_data %>% ggplot() + 
  aes(x = deltaBPM) + 
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Stim",
    y = "Density"
  )

# plotting the probabilities of answering 'faster'
hier_data %>%
  ggplot(aes(x = theta))  +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Probability of answering 'faster'",
    y = "Density"
  )


hier_data %>% 
  ggplot() +
  aes(deltaBPM, theta, color = n_agent) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  ) + coord_cartesian(xlim=c(-80,80)) # obs akser

```



```{r fig.width=10, fig.height=5}
# visualise simulated data (choices)

hier_data %>% filter(n_agent %in% 10:19) %>% 
  ggplot() +
  aes(deltaBPM, theta, color = as.factor(n_agent)) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  )  + facet_wrap(~n_agent) 


```


# Prior predictive checks - pre stan

```{r}
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, alpha) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = alpha)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
 # scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - alpha",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )


# BETA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, beta) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = beta)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 40)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - beta",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )



# LAMBDA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, lambda) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = lambda)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 0.1)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - lambda",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )
```

maybe this is not the best way to check prior predictive checks?


```{r}
# compile model
file = file.path(stan_model_dir, 'simple_model_hrd_hierarchical.stan')
model <- cmdstan_model(
  file)

# make data stan friendly i.e. in list form 
hier_stan_data <- list(
  N = nrow(hier_data),
  S = 60,
  subj = hier_data$n_agent,
  dBPM = hier_data$deltaBPM,
  choice = hier_data$choice,
  prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
  prior_mu_alpha_scale         = 7, #changed 6/7 from 7 to 15
  prior_sigma_alpha_loc        = 10,
  prior_sigma_alpha_scale      = 3,
  prior_mu_b_log_loc           = 2.5,    # exp(2.5) ~ 12 BPM slope
  prior_mu_b_log_scale         = 0.2,
  prior_sigma_b_log_loc        = 0.4,
  prior_sigma_b_log_scale      = 0.1,
  prior_mu_lambda_logit_loc    = -3,      # 0.5 * inv_logit(-3) ~ 0.024
  prior_mu_lambda_logit_scale  = 0.4,
  prior_sigma_lambda_logit_rate = 2.5,    # E[sigma_lambda_logit] = 0.4
  run_diagnostics = 0L # not needed now 
)

str(hier_stan_data)
```
3. Sample

```{r}
save_model = TRUE

# check if model has been run already
run_model <- regenerate_simulations || !file.exists(model_file)

if (!run_model) {
  fit <- readRDS(model_file)
}

if (run_model) {
  start_time <- Sys.time()

  pf <- model$pathfinder(data = hier_stan_data, num_paths = 10, draws = 400, psis_resample=FALSE)

  fit <- model$sample(
    data = hier_stan_data,
    init = pf,
    seed = 123,
    chains = 4,
    parallel_chains = 4,
    #threads_per_chain = 4,
    iter_warmup = 500,
    iter_sampling = 1000,
    refresh = 500#,
   # show_messages= FALSE
  )

  end_time <- Sys.time()
  time_equal <- end_time - start_time
  print(time_equal)
  
  if (save_model){
    fit$save_object(file = model_file)
  }
}
```

heck diagnostics

```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")

```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit$summary(variables = c("mu_alpha", "mu_b_log", "mu_lambda_logit"))
fit_summary %>% dplyr::select(variable, mean, rhat, ess_bulk, ess_tail)

draws_df <- as_draws_df(fit$draws())
```




```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha

# Drop variables that are all NA (run_diagnostics = 0 quantities)
draws_df_clean <- draws_df |>
  select(where(~ !all(is.na(.x))))

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_alpha", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```

```{r}
# 3. Visual Checks (Trace and Rank Plots) - beta

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_b_log") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_b_log", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_lambda_logit") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_lambda_logit", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```


```{r}
# then get generated quantities
fit_gq <- model$generate_quantities(
  fitted_params = fit,
  data = list(
    N = nrow(hier_data),
    S = 60,
    subj = hier_data$n_agent,
    dBPM = hier_data$deltaBPM,
    choice = hier_data$choice,
    prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
    prior_mu_alpha_scale         = 7, #changed 6/7 from 7 to 15
    prior_sigma_alpha_loc        = 10,
    prior_sigma_alpha_scale      = 3,
    prior_mu_b_log_loc           = 2.5,    # exp(2.5) ~ 12 BPM slope
    prior_mu_b_log_scale         = 0.2,
    prior_sigma_b_log_loc        = 0.4,
    prior_sigma_b_log_scale      = 0.1,
    prior_mu_lambda_logit_loc    = -3,      # 0.5 * inv_logit(-3) ~ 0.024
    prior_mu_lambda_logit_scale  = 0.4,
    prior_sigma_lambda_logit_rate = 2.5,    # E[sigma_lambda_logit] = 0.4
    run_diagnostics = 1L # NOW we want diagnostics
  )
)
```

#2. Prior predictive checks

Do the prior allow only plausible behavior?


```{r}
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block
draws_gq <- as_draws_df(fit_gq$draws()) # Ensure we have the draws as a dataframe
observed_choice <- sum(hier_data$choice) # Total number of 'right' choices in the actual observed data

ppc_prior_ggplot <- ggplot(draws_gq, aes(x = choice_prior_sum)) +
  # Draw the histogram of the model's prior predictions (light red)
  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  
  # Add the vertical line for the actual observed data (dark red)
  geom_vline(xintercept = observed_choice, color = "darkred", linewidth = 1.2) +
  
  # Labels and theming
  ggtitle("Prior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = histogram spans the full range; observed value (dashed) is not in the tails") +
  xlab(paste("Total Number of 'Faster' Choices (out of", nrow(hier_data), "trials)")) + 
  ylab("Frequency") +
  coord_cartesian(xlim = c(0,2400))

ppc_prior_ggplot
```
Central limit theorem - data will pile up around the center
observed choice sits in the bulk of the distribution
good: data spans almost entire possible range of data (from 0 to 2400)

```{r}
draws_gq <- as_draws_df(fit_gq$draws())

# Extract alpha_prior and beta_prior for all subjects across draws
# and compute implied p(faster) at dBPM = 0 for each subject
draws_gq |>
  select(starts_with("alpha_prior"), starts_with("beta_prior"), starts_with("lambda_prior")) |>
  pivot_longer(everything(), names_to = c(".value", "s"),
               names_pattern = "(alpha_prior|beta_prior|lambda_prior)\\[(\\d+)\\]") |>
  mutate(p_faster = lambda_prior + (1 - 2 * lambda_prior) * pnorm((0 - alpha_prior) / beta_prior)) |>
  ggplot(aes(x = p_faster)) +
  geom_histogram(bins = 50, fill = "#ffbaba", colour = "#ff5252") +
  xlab("Prior p(faster) at dBPM = 0") +
  ggtitle("Prior predictive: per-subject p(faster)",
          subtitle = "Should be roughly uniform across [0,1] for a weakly informative prior")



dBPM_grid <- unique(hier_data$deltaBPM)  # the actual stimulus values used

draws_gq |>
  select(starts_with("alpha_prior"), starts_with("beta_prior"), starts_with("lambda_prior")) |>
  pivot_longer(everything(), names_to = c(".value", "s"),
               names_pattern = "(alpha_prior|beta_prior|lambda_prior)\\[(\\d+)\\]") |>
  slice_sample(n = 200) |>  # subset for speed
  crossing(dBPM = dBPM_grid) |>
  mutate(p_faster = lambda_prior + (1 - 2 * lambda_prior) *
           pnorm((dBPM - alpha_prior) / beta_prior)) |>
  ggplot(aes(x = dBPM, y = p_faster, group = s)) +
  geom_line(alpha = 0.05, colour = "#ff5252") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  xlab("dBPM") + ylab("Prior p(faster)") +
  ggtitle("Prior predictive psychometric curves",
          subtitle = "Each line = one prior subject; should span a wide range of slopes and thresholds")
```

```{r}
draws_gq |>
  select(starts_with("alpha_prior")) |>
  pivot_longer(everything(), names_to = "subject", values_to = "alpha_prior") |>
  ggplot(aes(x = alpha_prior)) +
  geom_histogram(bins = 80, fill = "#ffbaba", colour = "#ff5252") +
  xlab("Prior alpha (BPM threshold)") +
  ggtitle("Prior predictive: per-subject threshold",
          subtitle = "Should cover roughly [-30, +10] BPM for HRD task")
```


```{r}
draws <- fit_gq$draws("theta_prior_p", format = "matrix")

# Option 1: histogram of ALL theta_prior_p values (collapsed across trials and draws)
hist(as.vector(draws), breaks = 50, 
     main = "Prior predictive theta", xlab = "theta", xlim = c(0,1))

# Option 2: plot the mean theta per trial with uncertainty
theta_mean <- apply(draws, 2, mean)
theta_lower <- apply(draws, 2, quantile, 0.05)
theta_upper <- apply(draws, 2, quantile, 0.95)

plot(hier_data$deltaBPM, theta_mean, ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive psychometric function")
segments(hier_data$deltaBPM, theta_lower, hier_data$deltaBPM, theta_upper, col = "grey")

# Option 3: plot a random sample of draws as psychometric curves (most informative)
n_curves <- 50
sampled <- sample(nrow(draws), n_curves)

plot(NULL, xlim = range(hier_data$deltaBPM), ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive curves")
for (i in sampled) {
  lines(sort(hier_data$deltaBPM), draws[i, order(hier_data$deltaBPM)], col = rgb(0,0,1,0.1))
}
```

#3. Posterior predictive checks

```{r}
n_trials = 60 # for plotting
# Extract posterior predictive simulations from the fitted object
# This matrix has dimensions: [number_of_draws, number_of_trials]
choice_pred <- fit_gq$draws("choice_pred", format = "matrix")


# Calculate the test statistic T(y) for the *actual* empirical data
observed_choice <- sum(hier_data$choice) 

# Calculate the test statistic T(y_rep) for *each* replicated dataset
# We compute the row-wise sum across the posterior draws
choice_pred_sum <- rowSums(choice_pred)

# Bundle into a tibble for strictly structured plotting
ppc_df <- tibble(
  Replicated_Sum = choice_pred_sum
)

# --- Graphical Posterior Predictive Check using ggplot2 ---
ppc_posterior_ggplot <- ggplot(ppc_df, aes(x = Replicated_Sum)) +
  geom_histogram(fill = "skyblue", color = "white", alpha = 0.7, bins = 30) +
  geom_vline(xintercept = observed_choice, color = "black", linetype = "dashed", linewidth = 1) +
  ggtitle("Posterior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = observed value (dashed) sits inside the bulk of replicated datasets") +
  xlab(paste("Total Number of 'Faster' Choices (out of", n_trials, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_choice, y = Inf, 
    label = "Sim data", 
    vjust = 2, hjust = if(observed_choice > n_trials/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) 
ppc_posterior_ggplot
```

```{r}
ppc_overlay_df <- tibble(
  Replicated_Sum = c(draws_gq$choice_prior_sum, choice_pred_sum),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(choice_pred_sum)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 2. Generate the overlay plot
ppc_overlay_ggplot <- ggplot(ppc_overlay_df, aes(x = Replicated_Sum, fill = Distribution)) +
  # Use position = "identity" and alpha to ensure both distributions are visible
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 30) +
  geom_vline(xintercept = observed_choice, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Prior vs. Posterior Predictive: Narrowing of Predictions After Fitting",
          subtitle = "Posterior (blue) should be much tighter than prior (red) and centred on observed data") +
  xlab(paste("Total Number of 'Right' Choices (out of", n_trials, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_choice, y = Inf, 
    label = "Empirical data", 
    vjust = 2, hjust = if(observed_choice > n_trials/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(ppc_overlay_ggplot)
```
```{r}
# --- Example of checking another statistic (e.g., max run length) ---
# Define a function to calculate the maximum run length in a vector
max_run_length <- function(x) {
  if (length(x) == 0 || all(is.na(x))) return(0) 
  rle_x <- rle(x) # Run Length Encoding calculates consecutive identical choices
  if (length(rle_x$lengths) == 0) return(0) 
  max(rle_x$lengths) 
}

# 2. Calculate the empirical test statistic T(y)
observed_max_run <- max_run_length(hier_data$choice)

# 3. Extract prior replicated matrices and compute T(y_rep) for the prior
# y_prior_rep is an S x N matrix generated from the prior distributions
y_prior_rep <- fit_gq$draws("choice_prior_pred", format = "matrix")
prior_max_runs <- apply(y_prior_rep, 1, max_run_length)

# 4. Extract posterior replicated matrices and compute T(y_rep) for the posterior
# y_post_rep is an S x N matrix generated from the posterior distributions
y_post_rep <- fit_gq$draws("choice_pred", format = "matrix")
post_max_runs <- apply(y_post_rep, 1, max_run_length)

# 5. Bundle into a strictly structured tibble for overlay plotting
ppc_run_df <- tibble(
  Replicated_Max_Run = c(prior_max_runs, post_max_runs),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(post_max_runs)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 6. Generate the overlay plot
p_ppc_max_run <- ggplot(ppc_run_df, aes(x = Replicated_Max_Run, fill = Distribution)) +
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 20) +
  geom_vline(xintercept = observed_max_run, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Posterior Predictive: Maximum Run Length of Consecutive Choices",
          subtitle = "Tests sequential structure — a model assuming i.i.d. trials should still pass this") +
  xlab("Test Statistic T(y_rep): Maximum Run Length of Consecutive Choices") +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_max_run, y = Inf, 
    label = "Empirical T(y)", 
    vjust = 2, hjust = -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(p_ppc_max_run)
```
```{r}
# 4a. Look at the Posterior Estimate - alpha
ggplot(draws_df) +
  geom_density(aes(mu_alpha, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(mu_alpha_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = hier_data$mu_alpha, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_alpha (threshold)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "alpha  (threshold)",
    y        = "Density",
    fill     = "Distribution"
  ) 
```
```{r}
# 4b. Look at the Posterior Estimate - beta - tror noget går galt her
ggplot(draws_df) +
  geom_density(aes(exp(mu_b_log), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(exp(mu_b_log_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = exp(hier_data$mu_b_log), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_beta (slope)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "beta (slope)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```

```{r}
# 4c. Look at the Posterior Estimate - lambda
ggplot(draws_df) +
  geom_density(aes(inv_logit(mu_lambda_logit), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(inv_logit(mu_lambda_logit_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = inv_logit(hier_data$mu_lambda_logit), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_lambda (lapserate)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "lambda (lapserate)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```
# 5. Sensitivity analysis
# first combine draws from fit without and with generated quantities and remove empty variables from first fit
```{r}
# combine draws
draws_priorsense <- posterior::bind_draws(
  fit$draws(variables = c("mu_alpha", "sigma_alpha",
                          "mu_b_log", "sigma_b_log",
                          "mu_lambda_logit", "sigma_lambda_logit")),
  fit_gq$draws(variables = c("lprior_mu_alpha", "lprior_sigma_alpha",
                              "lprior_mu_b_log", "lprior_sigma_b_log",
                              "lprior_mu_lambda_logit", "lprior_sigma_lambda_logit",
                              "log_lik"))
) |> posterior::as_draws_df()

# get specs
sensitivity_summary <- priorsense::powerscale_sensitivity(draws_priorsense,
    prior_selection = c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit"))
sensitivity_summary

# We extract the variables of interest
# param_metrics <- sensitivity_summary |>
#   dplyr::filter(variable %in% c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit"))

# cat("\n--- Epistemic Diagnosis ---\n")
# print(param_metrics)

# plot
p_priorsense_dens <- priorsense::powerscale_plot_dens(
  draws_priorsense,
  prior_selection = c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log",
                      "mu_lambda_logit", "sigma_lambda_logit"),
  variables = c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log",
                "mu_lambda_logit", "sigma_lambda_logit")
)
p_priorsense_dens
```
```{r}
# to get around memory issues
# draws_priorsense <- fit_gq$draws(
#   variables = c("mu_alpha", "sigma_alpha", 
#                 "lprior_mu_alpha", "lprior_sigma_alpha",
#                 "mu_b_log", "sigma_b_log",
#                 "lprior_mu_b_log", "lprior_sigma_b_log",
#                 "mu_lambda_logit", "sigma_lambda_logit",
#                 "lprior_mu_lambda_logit", "lprior_sigma_lambda_logit",
#                 "log_lik")
# ) |> posterior::as_draws_df()
# 
# p_priorsense_dens <- priorsense::powerscale_plot_dens(
#   draws_priorsense,
#   prior_selection = c("mu_alpha", "sigma_alpha","mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit"),
#   variables = c("mu_alpha", "sigma_alpha","mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit")
# )
# p_priorsense_dens
```
