---
title: "1_simple_hrd_model"
output: word_document
date: "2026-06-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# 1. Ensure 'pacman' is installed
if (!require("pacman")) install.packages("pacman")

# 2. Load required packages using pacman for efficient package management - ensures packages are installed if missing, then loads them
pacman::p_load(
  tidyverse,    # Core suite for data manipulation (dplyr) and plotting (ggplot2)
  here,         # For creating robust file paths relative to the project root (optional but good practice)
  posterior,    # Provides tools for working with posterior distributions
  cmdstanr,     # The R interface to Stan (via CmdStan)
  bayesplot,    # Specialized plotting functions for Bayesian models
  patchwork,    # For easily combining multiple ggplot plots
  priorsense,   # For automated prior sensitivity analysis
  SBC,          # Simulation-Based Calibration Checks
  future,       # Parallel processing
  furrr,         # Parallel mapping
  pracma        # to be used in the probit regression formula
)

# Set a default theme for ggplot for consistency
theme_set(theme_classic()) # Use a classic theme with white background

# Define relative paths 
stan_model_dir <- here("stan")
stan_results_dir <- here("simmodels")
sim_data_dir <- here("simdata")

# Create directories if they don't exist (important for first run)
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir)
if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)
if (!dir.exists(sim_data_dir)) dir.create(sim_data_dir)

# Set to TRUE to rerun everything (takes time!), FALSE to load saved results.
regenerate_simulations <- FALSE 
```

Clean up of code in "generating_data.Rmd" + "fit_to_simulated.Rmd"

#1. Mathematical formulation
Simple HRD model with lapse rate

$$

p(x \mid \alpha,\beta, \lambda) = \lambda + (1-2*\lambda)*(0.5+0.5*erf(\frac{x-\alpha}{\beta*\sqrt{2}}))

$$
which can be written in R as follows

```{r}
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
}
```

we also need the inverse logit function, to transform values from log-odds scale to probability scale (0-1)
```{r}
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```

## generating data

```{r}
alpha = -10
lambda = 0.05
beta = 10
N = 100

# new 
sim_data_sigrid = data.frame(stimulus = rnorm(N,alpha,2*beta), subject = 1) %>% mutate(trial = 1:n())

# generate probability of answering 1, faster based on psychometric function
sim_data_lapse = sim_data_sigrid %>% 
  select(c(subject, trial, stimulus))


# generate probability of answering 1, faster based on psychometric function
sim_data_lapse = sim_data_sigrid %>% 
  select(c(subject, trial, stimulus))

sim_data_lapse$p_faster <- probit_regression(
  x     = sim_data_lapse$stimulus,
  alpha = alpha,
  beta  = beta,
  lambda = lambda
)

# generate responses
sim_data_lapse$response <- rbinom(
  n = nrow(sim_data_lapse),
  size = 1,
  prob = sim_data_lapse$p_faster
)
```

## Plotting simulated data

```{r}
sim_data_lapse %>% ggplot() + 
  aes(x = stimulus) + 
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Stim",
    y = "Density"
  )

# plotting the probabilities of answering 'faster'
sim_data_lapse %>%
  ggplot(aes(x = p_faster))  +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Probability of answering 'faster'",
    y = "Density"
  )


sim_data_lapse %>%
  ggplot(aes(x = (stimulus), y = p_faster)) +
  geom_jitter()+
  labs(
    x = "Delta BPM",
    y = "Probability of answering 'faster'"
  )
```
## Getting into stan
```{r}
# make data stan friendly i.e. in list form 
simple_stan_data <- list(
  N = nrow(sim_data_lapse),
  dBPM = sim_data_lapse$stimulus,
  choice = sim_data_lapse$response
)

str(simple_stan_data)
```

```{r}
# specify stan file path and compile model 

file = file.path('simple_model_hrd.stan')
model <- cmdstan_model(
  file)#, 
#  cpp_options = list(stan_threads = FALSE), # enable threading - don't know what this means
#  stanc_options = list("O1")) # basic optimization - don't know how

```
```{r}
start_time <- Sys.time()

fit <- model$sample(
  data = simple_stan_data,
  seed = 123,
  chains = 4,
  parallel_chains = 4,
  #threads_per_chain = 4,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 500#,
 # show_messages= FALSE
)

end_time <- Sys.time()
time_equal <- end_time - start_time
print(time_equal)
```

```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")
```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit$summary(variables = c("alpha", "beta", "lambda"))
fit_summary %>% dplyr::select(variable, mean, rhat, ess_bulk, ess_tail)
```

```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "alpha", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - beta
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "beta") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "beta", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - lambda
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "lambda") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "lambda", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
Looks good.

#2. Prior predictive checks

Do the prior allow only plausible behavior?


```{r}
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block
draws_gq <- as_draws_df(fit$draws()) # Ensure we have the draws as a dataframe
observed_choice <- sum(sim_data_lapse$response) # Total number of 'right' choices in the actual observed data

ppc_prior_ggplot <- ggplot(draws_gq, aes(x = choice_prior_sum)) +
  # Draw the histogram of the model's prior predictions (light red)
  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  
  # Add the vertical line for the actual observed data (dark red)
  geom_vline(xintercept = observed_choice, color = "darkred", linewidth = 1.2) +
  
  # Labels and theming
  ggtitle("Prior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = histogram spans the full range; observed value (dashed) is not in the tails") +
  xlab(paste("Total Number of 'Faster' Choices (out of", N, "trials)")) +
  ylab("Frequency") + 
  coord_cartesian(xlim = c(0,100))

ppc_prior_ggplot
```
looks okay - rouhgly uniform distribution with a mild mound shape spanning the full trial range from 0 to 100, and the observed value (from the simulated data) is about in the middle - but skewed towards 'Faster'.

```{r}
# CmdStanR
draws <- fit$draws("theta_prior_p", format = "matrix")

# Option 1: histogram of ALL theta_prior_p values (collapsed across trials and draws)
hist(as.vector(draws), breaks = 50, 
     main = "Prior predictive theta", xlab = "theta", xlim = c(0,1))

# Option 2: plot the mean theta per trial with uncertainty
theta_mean <- apply(draws, 2, mean)
theta_lower <- apply(draws, 2, quantile, 0.05)
theta_upper <- apply(draws, 2, quantile, 0.95)

plot(sim_data_lapse$stimulus, theta_mean, ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive psychometric function")
segments(sim_data_lapse$stimulus, theta_lower, sim_data_lapse$stimulus, theta_upper, col = "grey")

# Option 3: plot a random sample of draws as psychometric curves (most informative)
n_curves <- 50
sampled <- sample(nrow(draws), n_curves)

plot(NULL, xlim = range(sim_data_lapse$stimulus), ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive curves")
for (i in sampled) {
  lines(sort(sim_data_lapse$stimulus), draws[i, order(sim_data_lapse$stimulus)], col = rgb(0,0,1,0.1))
}
```
#3. Posterior predictive checks

```{r}
# Extract posterior predictive simulations from the fitted object
# This matrix has dimensions: [number_of_draws, number_of_trials]
choice_pred <- fit$draws("choice_pred", format = "matrix")


# Calculate the test statistic T(y) for the *actual* empirical data
observed_choice <- sum(sim_data_lapse$response) 

# Calculate the test statistic T(y_rep) for *each* replicated dataset
# We compute the row-wise sum across the posterior draws
choice_pred_sums <- rowSums(choice_pred)

# Bundle into a tibble for strictly structured plotting
ppc_df <- tibble(
  Replicated_Sum = choice_pred_sums
)

# --- Graphical Posterior Predictive Check using ggplot2 ---
ppc_posterior_ggplot <- ggplot(ppc_df, aes(x = Replicated_Sum)) +
  geom_histogram(fill = "skyblue", color = "white", alpha = 0.7, bins = 30) +
  geom_vline(xintercept = observed_choice, color = "black", linetype = "dashed", linewidth = 1) +
  ggtitle("Posterior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = observed value (dashed) sits inside the bulk of replicated datasets") +
  xlab(paste("Total Number of 'Faster' Choices (out of", N, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_choice, y = Inf, 
    label = "Sim data", 
    vjust = 2, hjust = if(observed_choice > N/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) 
ppc_posterior_ggplot
```


#4. Prior posterior update checks

#5. Sensitivity analysis

#6. Parameter recovery and SBC
