---
title: "1_simple_hrd_model"
output: word_document
date: "2026-06-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# 1. Ensure 'pacman' is installed
if (!require("pacman")) install.packages("pacman")

# 2. Load required packages using pacman for efficient package management - ensures packages are installed if missing, then loads them
pacman::p_load(
  tidyverse,    # Core suite for data manipulation (dplyr) and plotting (ggplot2)
  here,         # For creating robust file paths relative to the project root (optional but good practice)
  posterior,    # Provides tools for working with posterior distributions
  cmdstanr,     # The R interface to Stan (via CmdStan)
  bayesplot,    # Specialized plotting functions for Bayesian models
  patchwork,    # For easily combining multiple ggplot plots
  priorsense,   # For automated prior sensitivity analysis
  SBC,          # Simulation-Based Calibration Checks
  future,       # Parallel processing
  furrr,        # Parallel mapping
  pracma,       # to be used in the probit regression formula
  cli,
  devtools,     # to download package from github
  progressr,     # to get progress bar in SBC
  remotes       # to install specific version of ggplot needed to make priorsense dens plot
)

devtools::install_github("hyunjimoon/SBC") # to make SBC
#progressr::handlers(global = TRUE) 
#progressr::handlers(c("rstudio", "progress")) # måske bedre??


# priorsense requires specific version of ggplot
#packageVersion("ggplot2")
install.packages("remotes")
remotes::install_version("ggplot2", version = "3.5.1")


# Set a default theme for ggplot for consistency
theme_set(theme_classic()) # Use a classic theme with white background

# Define relative paths 
stan_model_dir <- here("stan")
stan_results_dir <- here("simmodels")
sim_data_dir <- here("simdata")

# Create directories if they don't exist (important for first run)
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir)
if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)
if (!dir.exists(sim_data_dir)) dir.create(sim_data_dir)

# Define path for saving/loading the fitted model object that includes the generated quantities
model_file_gq <- file.path(stan_results_dir, "1_simple_hrd_model.rds")
sbc_path <- file.path(stan_results_dir, "1_simple_hrd_model_sbc.rds")

# Set to TRUE to rerun everything (takes time!), FALSE to load saved results.
regenerate_simulations <- FALSE 
simulate_more_SBC = TRUE
```

Clean up of code in "generating_data.Rmd" + "fit_to_simulated.Rmd" + parameter_recovery.Rmd

#1. Mathematical formulation
Simple HRD model with lapse rate

$$

p(x \mid \alpha,\beta, \lambda) = \lambda + (1-2*\lambda)*(0.5+0.5*erf(\frac{x-\alpha}{\beta*\sqrt{2}}))

$$
which can be written in R as follows

```{r}
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
}
```

we also need the inverse logit function, to transform values from log-odds scale to probability scale (0-1)
```{r}
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```

## generating data

```{r}
alpha = -10
lambda = 0.05
beta = 10
N = 100

# new 
sim_data_sigrid = data.frame(stimulus = rnorm(N,alpha,2*beta), subject = 1) %>% mutate(trial = 1:n())

# generate probability of answering 1, faster based on psychometric function
sim_data_lapse = sim_data_sigrid %>% 
  select(c(subject, trial, stimulus))


# generate probability of answering 1, faster based on psychometric function
sim_data_lapse = sim_data_sigrid %>% 
  select(c(subject, trial, stimulus))

sim_data_lapse$p_faster <- probit_regression(
  x     = sim_data_lapse$stimulus,
  alpha = alpha,
  beta  = beta,
  lambda = lambda
)

# generate responses
sim_data_lapse$response <- rbinom(
  n = nrow(sim_data_lapse),
  size = 1,
  prob = sim_data_lapse$p_faster
)
```

## Plotting simulated data

```{r}
sim_data_lapse %>% ggplot() + 
  aes(x = stimulus) + 
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Stim",
    y = "Density"
  )

# plotting the probabilities of answering 'faster'
sim_data_lapse %>%
  ggplot(aes(x = p_faster))  +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Probability of answering 'faster'",
    y = "Density"
  )


sim_data_lapse %>%
  ggplot(aes(x = (stimulus), y = p_faster)) +
  geom_jitter()+
  labs(
    x = "Delta BPM",
    y = "Probability of answering 'faster'"
  )
```
## Getting into stan
```{r}
# make data stan friendly i.e. in list form 
simple_stan_data <- list(
  N = nrow(sim_data_lapse),
  dBPM = sim_data_lapse$stimulus,
  choice = sim_data_lapse$response,
  run_diagnostics = 1L # single model check
)

str(simple_stan_data)
```

```{r}
# specify stan file path and compile model 
file = file.path(stan_model_dir, 'simple_model_hrd.stan')
model_gq <- cmdstan_model(
  file)#, 
#  cpp_options = list(stan_threads = FALSE), # enable threading - don't know what this means
#  stanc_options = list("O1")) # basic optimization - don't know how


# check if model has been run already
run_model <- regenerate_simulations || !file.exists(model_file_gq)

if (!run_model) {
  fit_gq <- readRDS(model_file_gq)
}

if (run_model) {
  start_time <- Sys.time()

  fit_gq <- model_gq$sample(
    data = simple_stan_data,
    seed = 123,
    chains = 4,
    parallel_chains = 4,
    #threads_per_chain = 4,
    iter_warmup = 1000,
    iter_sampling = 2000, # 8000 posterior draws in total
    refresh = 500#,
   # show_messages= FALSE
  )

  end_time <- Sys.time()
  time_equal <- end_time - start_time
  print(time_equal)
  
  fit_gq$save_object(file = model_file_gq)
}
```


```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit_gq$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")
```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit_gq$summary(variables = c("alpha", "beta", "lambda"))
fit_summary %>% dplyr::select(variable, mean, rhat, ess_bulk, ess_tail)
```

```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit_gq$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "alpha", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - beta
draws_df <- as_draws_df(fit_gq$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "beta") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "beta", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - lambda
draws_df <- as_draws_df(fit_gq$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "lambda") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "lambda", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
Looks good.

#2. Prior predictive checks

Do the prior allow only plausible behavior?


```{r}
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block
draws_gq <- as_draws_df(fit_gq$draws()) # Ensure we have the draws as a dataframe
observed_choice <- sum(sim_data_lapse$response) # Total number of 'right' choices in the actual observed data

ppc_prior_ggplot <- ggplot(draws_gq, aes(x = choice_prior_sum)) +
  # Draw the histogram of the model's prior predictions (light red)
  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  
  # Add the vertical line for the actual observed data (dark red)
  geom_vline(xintercept = observed_choice, color = "darkred", linewidth = 1.2) +
  
  # Labels and theming
  ggtitle("Prior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = histogram spans the full range; observed value (dashed) is not in the tails") +
  xlab(paste("Total Number of 'Faster' Choices (out of", N, "trials)")) +
  ylab("Frequency") #+ 
  coord_cartesian(xlim = c(0,100))

ppc_prior_ggplot
```
looks okay - rouhgly uniform distribution with a mild mound shape spanning the full trial range from 0 to 100, and the observed value (from the simulated data) is about in the middle - but skewed towards 'Faster'.

```{r}
# CmdStanR
draws <- fit_gq$draws("theta_prior_p", format = "matrix")

# Option 1: histogram of ALL theta_prior_p values (collapsed across trials and draws)
hist(as.vector(draws), breaks = 50, 
     main = "Prior predictive theta", xlab = "theta", xlim = c(0,1))

# Option 2: plot the mean theta per trial with uncertainty
theta_mean <- apply(draws, 2, mean)
theta_lower <- apply(draws, 2, quantile, 0.05)
theta_upper <- apply(draws, 2, quantile, 0.95)

plot(sim_data_lapse$stimulus, theta_mean, ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive psychometric function")
segments(sim_data_lapse$stimulus, theta_lower, sim_data_lapse$stimulus, theta_upper, col = "grey")

# Option 3: plot a random sample of draws as psychometric curves (most informative)
n_curves <- 50
sampled <- sample(nrow(draws), n_curves)

plot(NULL, xlim = range(sim_data_lapse$stimulus), ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive curves")
for (i in sampled) {
  lines(sort(sim_data_lapse$stimulus), draws[i, order(sim_data_lapse$stimulus)], col = rgb(0,0,1,0.1))
}
```
First plots shows the problem which the ordered beta regression aims to solve exactly! More on that in another model.

#3. Posterior predictive checks

```{r}
# Extract posterior predictive simulations from the fitted object
# This matrix has dimensions: [number_of_draws, number_of_trials]
choice_pred <- fit_gq$draws("choice_pred", format = "matrix")


# Calculate the test statistic T(y) for the *actual* empirical data
observed_choice <- sum(sim_data_lapse$response) 

# Calculate the test statistic T(y_rep) for *each* replicated dataset
# We compute the row-wise sum across the posterior draws
choice_pred_sum <- rowSums(choice_pred)

# Bundle into a tibble for strictly structured plotting
ppc_df <- tibble(
  Replicated_Sum = choice_pred_sum
)

# --- Graphical Posterior Predictive Check using ggplot2 ---
ppc_posterior_ggplot <- ggplot(ppc_df, aes(x = Replicated_Sum)) +
  geom_histogram(fill = "skyblue", color = "white", alpha = 0.7, bins = 30) +
  geom_vline(xintercept = observed_choice, color = "black", linetype = "dashed", linewidth = 1) +
  ggtitle("Posterior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = observed value (dashed) sits inside the bulk of replicated datasets") +
  xlab(paste("Total Number of 'Faster' Choices (out of", N, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_choice, y = Inf, 
    label = "Sim data", 
    vjust = 2, hjust = if(observed_choice > N/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) 
ppc_posterior_ggplot
```
So what we see here is on the x-axis the number of trials (0-100) i.e. the maximum number of 'faster' choices (if one choses 'faster' (1) on all 100 trials, the data point of the sum would be 100) and on the y-axis how many of the posterior samples (of the simulated datasets) have that number of 'faster' choices in their dataset.

This looks good - data is roughly centered around the middle but skewed towards choosing 'faster' more often than chance because the model and the emp data has a negative bias in the threshold, meaning that participants often reply 'faster' before the sounds are actually faster. 

```{r}
ppc_overlay_df <- tibble(
  Replicated_Sum = c(draws_gq$choice_prior_sum, choice_pred_sum),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(choice_pred_sum)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 2. Generate the overlay plot
ppc_overlay_ggplot <- ggplot(ppc_overlay_df, aes(x = Replicated_Sum, fill = Distribution)) +
  # Use position = "identity" and alpha to ensure both distributions are visible
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 30) +
  geom_vline(xintercept = observed_choice, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Prior vs. Posterior Predictive: Narrowing of Predictions After Fitting",
          subtitle = "Posterior (blue) should be much tighter than prior (red) and centred on observed data") +
  xlab(paste("Total Number of 'Right' Choices (out of", N, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_choice, y = Inf, 
    label = "Empirical data", 
    vjust = 2, hjust = if(observed_choice > N/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(ppc_overlay_ggplot)
```
```{r}
# --- Example of checking another statistic (e.g., max run length) ---
# Define a function to calculate the maximum run length in a vector
max_run_length <- function(x) {
  if (length(x) == 0 || all(is.na(x))) return(0) 
  rle_x <- rle(x) # Run Length Encoding calculates consecutive identical choices
  if (length(rle_x$lengths) == 0) return(0) 
  max(rle_x$lengths) 
}

# 2. Calculate the empirical test statistic T(y)
observed_max_run <- max_run_length(sim_data_lapse$response)

# 3. Extract prior replicated matrices and compute T(y_rep) for the prior
# y_prior_rep is an S x N matrix generated from the prior distributions
y_prior_rep <- fit_gq$draws("choice_prior_pred", format = "matrix")
prior_max_runs <- apply(y_prior_rep, 1, max_run_length)

# 4. Extract posterior replicated matrices and compute T(y_rep) for the posterior
# y_post_rep is an S x N matrix generated from the posterior distributions
y_post_rep <- fit_gq$draws("choice_pred", format = "matrix")
post_max_runs <- apply(y_post_rep, 1, max_run_length)

# 5. Bundle into a strictly structured tibble for overlay plotting
ppc_run_df <- tibble(
  Replicated_Max_Run = c(prior_max_runs, post_max_runs),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(post_max_runs)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 6. Generate the overlay plot
p_ppc_max_run <- ggplot(ppc_run_df, aes(x = Replicated_Max_Run, fill = Distribution)) +
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 20) +
  geom_vline(xintercept = observed_max_run, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Posterior Predictive: Maximum Run Length of Consecutive Choices",
          subtitle = "Tests sequential structure — a model assuming i.i.d. trials should still pass this") +
  xlab("Test Statistic T(y_rep): Maximum Run Length of Consecutive Choices") +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_max_run, y = Inf, 
    label = "Empirical T(y)", 
    vjust = 2, hjust = -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(p_ppc_max_run)
```
To test if the trials are independent - which our model assumes. It looks good, because the model doesn't have very long runs of choosing the same option again and again.

#4. Prior posterior update checks

```{r}
# 4a. Look at the Posterior Estimate - alpha
ggplot(draws_df) +
  geom_density(aes(alpha, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(alpha_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = alpha, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: Alpha (threshold)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "alpha  (threshold)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```

```{r}
# 4b. Look at the Posterior Estimate - beta
ggplot(draws_df) +
  geom_density(aes(beta, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(beta_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = beta, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: Beta (slope)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "beta (slope)",
    y        = "Density",
    fill     = "Distribution"
  ) 
```
```{r}
# 4c. Look at the Posterior Estimate - lambda
ggplot(draws_df) +
  geom_density(aes(lambda, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(lambda_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = lambda, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: Lambda (lapserate)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "lambda (lapserate)",
    y        = "Density",
    fill     = "Distribution"
  ) 

```
Lambda doesn't look too good.

#5. Sensitivity analysis

```{r}
# We run prior sense
sensitivity_summary <- priorsense::powerscale_sensitivity(fit_gq,
    prior_selection = c("alpha", "b_log", "lambda_logit"))
sensitivity_summary

# We extract the variables of interest
param_metrics <- sensitivity_summary |>
  dplyr::filter(variable %in% c("alpha", "b_log", "lambda_logit"))

cat("\n--- Epistemic Diagnosis ---\n")
print(param_metrics)

# Visualize the geometric deformation of the posterior
p_priorsense_dens <- priorsense::powerscale_plot_dens(
  fit_gq,
  prior_selection = c("alpha", "b_log", "lambda_logit"),
  variables = c("alpha", "b_log", "lambda_logit")
) 

p_priorsense_dens

```
```{r}
print(param_metrics)
```
Alpha looks good, beta is okay, and lambda is not looking good.

#6. Parameter recovery and SBC

```{r}
# Define the model
file = file.path(stan_model_dir, 'simple_model_hrd.stan')
model <- cmdstan_model(file)

# Define the backend engine
backend <- SBC::SBC_backend_cmdstan_sample(
  model,
  chains = 4, # changed from 4 to 1
  parallel_chains = 4, # changed from 4 to 1 because I am handling parallellism with future
  iter_warmup = 1000,
  iter_sampling = 1000,
  refresh = 100
)

# --- GROUP SETTINGS ---
S <- 60      # number of participants per group dataset
N <- 100     # trials per participant
#n_trials_min = 40
#n_trials_max = 100
# total trials per dataset = S * N

# Define the generator
#generate_group_dataset <- function(S, n_trials_min, n_trials_max){
generate_group_dataset <- function(S, N){

  #parameters shared across all participants
  alpha_true = rnorm(n = 1, mean = -10, sd = 10) # kør bredere SD
  b_log_true = rnorm(n = 1, mean = 2.5, sd = 0.6)
  beta_true = exp(b_log_true)
  lambda_logit_true = rnorm(n = 1, mean = -3, sd = 0.4)
  lambda_true= 0.5*inv_logit(lambda_logit_true)
  
  # storage
  dBPM_all <- numeric(S*N)
  choice_all <- integer(S*N)

  idx <- 1

  for (s in 1:S) {
    # stimuli per participant
    deltaBPM <- rnorm(N, alpha_true, 2*beta_true)

    # choice probability
    theta_prob <- probit_regression(
      x = deltaBPM,
      alpha = alpha_true,
      beta = beta_true,
      lambda = lambda_true
    )

    # simulate binary choices
    y_sim <- rbinom(N, size = 1, prob = theta_prob)

    # place in joint data vector
    dBPM_all[idx:(idx+N-1)] <- deltaBPM
    choice_all[idx:(idx+N-1)] <- y_sim
    idx <- idx + N
  }

  list(
    variables = list(
      alpha = alpha_true,
      b_log = b_log_true,
      lambda_logit = lambda_logit_true
    ),
    generated = list(
      N = S*N,
      dBPM = dBPM_all,
      choice = choice_all,
      run_diagnostics = 0L # SBC doesn't need all generated quantities
    )
  )
}


# Register generator
generator <- SBC::SBC_generator_function(
  generate_group_dataset,
  S = S,
  N = N
)

# Run SBC - see next chunk
#datasets <- SBC::generate_datasets(generator, n_sims = 100)
#sbc_results <- SBC::compute_SBC(datasets, backend, keep_fits = FALSE)
```

# check if code below works
100 sims tog >40 min
```{r}
if (simulate_more_SBC || !file.exists(sbc_path)) {
 # future::plan(future::multisession, workers = 8) # I have 12  - dangerous!
  sbc_datasets_simple <- SBC::generate_datasets(generator, n_sims = 50)
  sbc_results_simple <- SBC::compute_SBC(sbc_datasets_simple, backend, keep_fits = FALSE)

  if (file.exists(sbc_path)){
    #load existing simulations
    existing_sims <- readRDS(sbc_path)
    # append new ones to existing
    combined_datasets <- SBC::bind_datasets(existing_sims$datasets, sbc_datasets_simple)
    combined_results <- SBC::bind_results(existing_sims$results, sbc_results_simple)
    # save everything
    saveRDS(
      list(datasets = combined_datasets, results = combined_results),
      sbc_path
    )
  }
  
  if (!file.exists(sbc_path)){
    saveRDS(
      list(datasets = sbc_datasets_simple, results = sbc_results_simple),
      sbc_path
    )
  }

} else {
  sbc_obj        <- readRDS(sbc_path)
  sbc_datasets_simple <- sbc_obj$datasets
  sbc_results_simple  <- sbc_obj$results
}
```

```{r}
# sanity check
nrow(sbc_obj$results$stats)
nrow(sbc_results_simple$stats)
nrow(combined_results$stats)

```

```{r}
# might have to reload data
sbc_obj        <- readRDS(sbc_path)
sbc_datasets_simple <- sbc_obj$datasets
sbc_results_simple  <- sbc_obj$results
  
# check diagnostics
sbc_results_simple
sbc_results_simple$backend_diagnostics
hist(sbc_results_simple$stats$rhat)
```


```{r} 
# number of simulations
n_sims = nrow(sbc_results_simple$stats)/3 # number of variables - check by writing #sbc_datasets_simple$variables

# parameter recovery
SBC::plot_sim_estimated(sbc_results_simple, variables = c("alpha", "b_log", "lambda_logit")) +
  labs(
    title    = sprintf("Parameter recovery with %d simulations", n_sims),
    subtitle = "Healthy = simulated values correlate with actual value"
  )
```


```{r}
SBC::plot_rank_hist(sbc_results_simple)

# with titles
plot_rank_hist(sbc_results_simple, variables = c("alpha", "b_log", "lambda_logit")) +
  labs(
    title    = sprintf("SBC Rank Histograms with %d simulations", n_sims),
    subtitle = "Healthy = bars flat within grey band. U-shape = overconfident."
  )

 SBC::plot_ecdf(sbc_results_simple, variables = c("alpha", "b_log", "lambda_logit")) +
  ggtitle("SBC ECDF: Deviation from Uniformity",
          subtitle = "Healthy = blue line stays inside the grey ellipse at all points") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "gray95", color = "NA"))
 
 plot_ecdf_diff(sbc_results_simple, vvariables = c("alpha", "b_log", "lambda_logit")) +
  labs(
    title    = "SBC ECDF Difference",
    subtitle = "Healthy = blue line stays near zero. Excursions here correlate with the divergence failures shown below."
  )
```

```{r}
# Extract the lightweight stats dataframe — already computed as part of SBC above.
sbc_stats <- sbc_results_simple$stats

# We plot on the probability scale for interpretability, matching Phase 6a.
# The 'theta' variable was stored (on probability scale) alongside 'theta_logit'
# in our generator's variables list.
recovery_sbc_df <- sbc_stats |>
  dplyr::filter(variable == "alpha")

# ---------------------------------------------------------------------------
# Identity plot: true value (x) vs. posterior median (y)
# ---------------------------------------------------------------------------
# Under perfect recovery, every point lies on the dashed y = x diagonal.
# Scatter around the diagonal is unavoidable estimation uncertainty.
# Systematic curvature (bowing above or below) would reveal directional bias.
# ---------------------------------------------------------------------------
p_sbc_recovery <- ggplot(recovery_sbc_df, aes(x = simulated_value, y = median)) +

  # Perfect-recovery reference line
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = "gray50", linewidth = 0.8) +

  # 90% posterior credible intervals
  # Under correct calibration, ~90% of bars should contain the true value —
  # which we already verified via the rank histogram above.
  geom_errorbar(aes(ymin = q5, ymax = q95),
                width = 0, alpha = 0.25, color = "steelblue") +

  # # Posterior median point estimates
  geom_point(aes(median), color = "midnightblue", alpha = 0.6, size = 1) +

 # coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +

  labs(
    title    = "Parameter Recovery via SBC: Simple model",
    subtitle = paste0(
      "N = 1000 simulated universes. True alpha drawn from the prior predictive.\n",
      "Each point is one universe. Bars are 90% posterior credible intervals."
    ),
    x = expression("True " * alpha * " (drawn from prior)"),
    y = "Posterior Median"
  ) +
  theme_classic()

print(p_sbc_recovery)

# ---------------------------------------------------------------------------
# Quantitative recovery metrics
# ---------------------------------------------------------------------------
# Pearson r   : linear correlation between truth and estimate (1.0 = perfect)
# RMSE        : average estimation error on the probability scale (0 = perfect)
# coverage90  : empirical 90% CI coverage — should be near 0.90 under calibration
# ---------------------------------------------------------------------------

recovery_metrics_sbc <- recovery_sbc_df |>
  dplyr::summarize(
    Pearson_r  = cor(simulated_value, median),
    RMSE       = sqrt(mean((simulated_value - median)^2)),
    coverage90 = mean(simulated_value >= q5 & simulated_value <= q95)
  )

cat("=== SBC PARAMETER RECOVERY METRICS (alpha) ===\n")
recovery_metrics_sbc
```
```{r}
recovery_sbc_df <- sbc_stats |>
  dplyr::filter(variable == "b_log")

# ---------------------------------------------------------------------------
# Identity plot: true value (x) vs. posterior median (y)
# ---------------------------------------------------------------------------
# Under perfect recovery, every point lies on the dashed y = x diagonal.
# Scatter around the diagonal is unavoidable estimation uncertainty.
# Systematic curvature (bowing above or below) would reveal directional bias.
# ---------------------------------------------------------------------------
p_sbc_recovery <- ggplot(recovery_sbc_df, aes(x = simulated_value, y = median)) +

  # Perfect-recovery reference line
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = "gray50", linewidth = 0.8) +

  # 90% posterior credible intervals
  # Under correct calibration, ~90% of bars should contain the true value —
  # which we already verified via the rank histogram above.
  geom_errorbar(aes(ymin = q5, ymax = q95),
                width = 0, alpha = 0.25, color = "steelblue") +

  # # Posterior median point estimates
  geom_point(aes(median), color = "midnightblue", alpha = 0.6, size = 1) +

 # coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +

  labs(
    title    = "Parameter Recovery via SBC: Simple model",
    subtitle = paste0(expression(
      "N = 1000 simulated universes. True " *beta * "_log drawn from the prior predictive.\n",
      "Each point is one universe. Bars are 90% posterior credible intervals."
    )),
    x = expression("True " * beta * "_log (drawn from prior)"),
    y = "Posterior Median"
  ) +
  theme_classic()

print(p_sbc_recovery)

# ---------------------------------------------------------------------------
# Quantitative recovery metrics
# ---------------------------------------------------------------------------
# Pearson r   : linear correlation between truth and estimate (1.0 = perfect)
# RMSE        : average estimation error on the probability scale (0 = perfect)
# coverage90  : empirical 90% CI coverage — should be near 0.90 under calibration
# ---------------------------------------------------------------------------

recovery_metrics_sbc <- recovery_sbc_df |>
  dplyr::summarize(
    Pearson_r  = cor(simulated_value, median),
    RMSE       = sqrt(mean((simulated_value - median)^2)),
    coverage90 = mean(simulated_value >= q5 & simulated_value <= q95)
  )

cat("=== SBC PARAMETER RECOVERY METRICS (beta, log scale) ===\n")
recovery_metrics_sbc
```

```{r}
recovery_sbc_df <- sbc_stats |>
  dplyr::filter(variable == "lambda_logit")

# ---------------------------------------------------------------------------
# Identity plot: true value (x) vs. posterior median (y)
# ---------------------------------------------------------------------------
# Under perfect recovery, every point lies on the dashed y = x diagonal.
# Scatter around the diagonal is unavoidable estimation uncertainty.
# Systematic curvature (bowing above or below) would reveal directional bias.
# ---------------------------------------------------------------------------
p_sbc_recovery <- ggplot(recovery_sbc_df, aes(x = simulated_value, y = median)) +

  # Perfect-recovery reference line
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = "gray50", linewidth = 0.8) +

  # 90% posterior credible intervals
  # Under correct calibration, ~90% of bars should contain the true value —
  # which we already verified via the rank histogram above.
  geom_errorbar(aes(ymin = q5, ymax = q95),
                width = 0, alpha = 0.25, color = "steelblue") +

  # # Posterior median point estimates
  geom_point(aes(median), color = "midnightblue", alpha = 0.6, size = 1) +

 # coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +

  labs(
    title    = "Parameter Recovery via SBC: Simple model",
    subtitle = paste0(expression(
      "N = 1000 simulated universes. True " *lambda * "_logit drawn from the prior predictive.\n",
      "Each point is one universe. Bars are 90% posterior credible intervals."
    )),
    x = expression("True " *lambda * "_logit (drawn from prior)"),
    y = "Posterior Median"
  ) +
  theme_classic()

print(p_sbc_recovery)

# ---------------------------------------------------------------------------
# Quantitative recovery metrics
# ---------------------------------------------------------------------------
# Pearson r   : linear correlation between truth and estimate (1.0 = perfect)
# RMSE        : average estimation error on the probability scale (0 = perfect)
# coverage90  : empirical 90% CI coverage — should be near 0.90 under calibration
# ---------------------------------------------------------------------------

recovery_metrics_sbc <- recovery_sbc_df |>
  dplyr::summarize(
    Pearson_r  = cor(simulated_value, median),
    RMSE       = sqrt(mean((simulated_value - median)^2)),
    coverage90 = mean(simulated_value >= q5 & simulated_value <= q95)
  )

cat("=== SBC PARAMETER RECOVERY METRICS (lambda, logit scale) ===\n")
recovery_metrics_sbc
```
```{r}

```


```{r}
# Extract diagnostics now so they are available for the targeted refit below
# diags_cp <- sbc_results_simple$backend_diagnostics
# cp_fail_rate <- mean(diags_cp$n_divergent > 0) * 100
# 
# # Divergence count distribution across all 200 universes
# diags_cp |> 
#   ggplot(aes(x = n_divergent)) +
#   geom_histogram(binwidth = 1, fill = "darkred", alpha = 0.8) +
#   labs(
#     title = "CP Divergence Counts Across 1000 SBC Universes",
#     subtitle = sprintf("%.0f%% of universes produced at least one divergence. 
# Even a single divergence invalidates the geometric guarantee of NUTS.", cp_fail_rate),
#     x = "Number of Divergent Transitions",
#     y = "Number of SBC Universes"
#   )
```

