---
title: "hierarchical model validation"
author: "Sigrid Bom"
date: "2026-07-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
# 1. Ensure 'pacman' is installed
if (!require("pacman")) install.packages("pacman")

# 2. Load required packages using pacman for efficient package management - ensures packages are installed if missing, then loads them
pacman::p_load(
  tidyverse,    # Core suite for data manipulation (dplyr) and plotting (ggplot2)
  here,         # For creating robust file paths relative to the project root (optional but good practice)
  posterior,    # Provides tools for working with posterior distributions
  cmdstanr,     # The R interface to Stan (via CmdStan)
  bayesplot,    # Specialized plotting functions for Bayesian models
  patchwork,    # For easily combining multiple ggplot plots
  priorsense,   # For automated prior sensitivity analysis
  SBC,          # Simulation-Based Calibration Checks
  future,       # Parallel processing
  furrr,        # Parallel mapping
  pracma,       # to be used in the probit regression formula
  cli,
  devtools,     # to download package from github
  progressr,     # to get progress bar in SBC
  remotes,       # to install specific version of ggplot needed to make priorsense dens plot
  msm
)

devtools::install_github("hyunjimoon/SBC") # to make SBC
#progressr::handlers(global = TRUE) 
#progressr::handlers(c("rstudio", "progress")) # måske bedre??


# priorsense requires specific version of ggplot
#packageVersion("ggplot2")
install.packages("remotes")
remotes::install_version("ggplot2", version = "3.5.1")


# Set a default theme for ggplot for consistency
theme_set(theme_classic()) # Use a classic theme with white background

# Define relative paths 
stan_model_dir <- here("stan")
stan_results_dir <- here("simmodels")
sim_data_dir <- here("simdata")

# Create directories if they don't exist (important for first run)
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir)
if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)
if (!dir.exists(sim_data_dir)) dir.create(sim_data_dir)

# Define path for saving/loading the fitted model object that includes the generated quantities
model_file <- file.path(stan_results_dir, "2_simple_hrd_model_hier.rds")
sbc_path <- file.path(stan_results_dir, "2_simple_hrd_model_hier_sbc.rds")

# Set to TRUE to rerun everything (takes time!), FALSE to load saved results.
regenerate_simulations <- FALSE 
simulate_more_SBC = TRUE
```
Clean up of code in "hierarchical_hrd_model_sim_and_test.Rmd"


# Mathematical formulation

Each agent j {1, ..., J} has an unobserved parameter \theta_j on each trial i, which decides the choice _h_ of agent j on each trial i.

The theta parameter is found through a probit regression, which depends on several parameters including the stimulus level (delta BPM, x), the threshold (alpha), the slope (beta) and the lapse rate (lambda). All the parameters which are used to calculate theta are population level parameters, which means that they are drawn from a distrubition with a mean (mu) and standard deviation (sigma).

$$
h_i,_j \sim Bernoulli (\theta_i) \\
 \theta_i = \lambda + (1-2*\lambda)*(0.5+0.5*erf(\frac{x-\alpha}{\beta*\sqrt{2}}))
$$
The latent parameters are thought to be distributed as follows:
$$
\alpha_j \sim Normal(\mu_\alpha, \sigma_\alpha) \\
\beta log_j \sim Normal(\mu_\beta, \sigma_\beta) \\
\lambda  logit_j \sim Normal (\mu_\lambda, \sigma_\lambda)
$$
Overview of hyperparameters:
$$
\mu_\alpha \sim Normal(-10,7) \\
\sigma_\alpha \sim Exponential(10) \\

\mu_\beta log \sim Normal(2.5,0.4) \\
\sigma_\beta log \sim Exponential(0.4)\\

\mu_\lambda logit \sim Normal(-3,0.4) \\
\sigma_\lambda logit \sim Exponential(0.4) \\
$$

Because beta needs to be positive, we sample it in the logarithmic space and then transform it back after sampling by exponentiating it. 
Similarly, we want lambda to be a small positive number, so we take the inverse logit of lambda and multiply it by 0.5.

Transformed parameters:
$$
\beta_j = exp(\beta log_j) \\
\lambda_j = 0.5* \frac{1}{(1+exp(-\lambda logit_j))}
$$

```{r}
# define functions

# probit regression - psychometric function
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
 }

# inverse logit
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```


# 1. Generating model

```{r}
simulate_hierarchical_hrd_data <- function(n_agent = 60,
                                           n_trials = 40,
                                           mu_alpha = NULL,
                                           sigma_alpha = NULL,
                                           mu_b_log = NULL,
                                           sigma_b_log = NULL,
                                           mu_lambda_logit = NULL,
                                           sigma_lambda_logit = NULL,
                                           seed = NULL
                                           ){
 
  if (!is.null(seed)) set.seed(seed)    
  
  # 1. sample population hyperparameters (if not provided)
  if (is.null(mu_alpha)) mu_alpha <- rnorm(1, mean = -10, sd = 15) # skal outputte omkring -10  # ændret fra 7 til 15
  if (is.null(sigma_alpha)) sigma_alpha <- rexp(1, rate = 1/10) # skal outputte omkring 10
  
  if (is.null(mu_b_log)) mu_b_log <- rnorm(1, mean = 2.5, sd = 0.4) # skal outputte omkring 2.5 
  if (is.null(sigma_b_log)) sigma_b_log <- rexp(1, rate = 1/0.4) # skal outputte omrking 0.4 
  
  if (is.null(mu_lambda_logit)) mu_lambda_logit <- rnorm(1, mean = -3, sd = 0.4) # skal være omrking -3
  if (is.null(sigma_lambda_logit)) sigma_lambda_logit <- rexp(1, rate = 1/0.4) #skal outputte omkring 0.4
  
  # 1.5 sample individual parameter values using hyper parameter values from population level
  
  data_sim = map_dfr(1:n_agent, ~ {
    # 1. sample true parameters per agent
    alpha         <- rnorm(1, mu_alpha, sigma_alpha)
    b_log         <- rnorm(1, mu_b_log, sigma_b_log)
    beta          <- exp(b_log)
    lambda_logit <- rnorm(1, mu_lambda_logit, sigma_lambda_logit)
    lambda       <- 0.5 * inv_logit(lambda_logit)
  
    # 2. sample stim
    deltaBPM <- msm::rtnorm(n_trials, mean = alpha, sd = 2 * beta, lower = -50, upper = 50) # truncated normal # is it a bad thing to depend the stimuli on the sampled parameters? also should this be different for each participant like it is now?
  
    # 3. calculate theta and choice
    theta_list <- probit_regression(x = deltaBPM, alpha = alpha, beta = beta, lambda = lambda)
    choice = rbinom(n=n_trials, size = 1, prob = theta_list) 
  
    # 4. return a tibble for each agent
    tibble(
      n_agent   = .x,
      n_trials,
      alpha,
      mu_alpha,
      sigma_alpha,
      beta,
      mu_b_log,
      sigma_b_log,
      lambda,
      mu_lambda_logit,
      sigma_lambda_logit,
      deltaBPM,
      theta = theta_list,
      choice
    )
})
  
 return(data_sim) 
}

```


```{r}

# need to set seed to get same response every time

hier_data = simulate_hierarchical_hrd_data(seed = 123) 

hier_data
```

```{r}
hier_data %>% ggplot() + 
  aes(x = deltaBPM) + 
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Stim",
    y = "Density"
  )

# plotting the probabilities of answering 'faster'
hier_data %>%
  ggplot(aes(x = theta))  +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Probability of answering 'faster'",
    y = "Density"
  )


hier_data %>% 
  ggplot() +
  aes(deltaBPM, theta, color = n_agent) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  ) + coord_cartesian(xlim=c(-80,80)) # obs akser

```


```{r fig.width=10, fig.height=5}
# visualise simulated data (choices)

hier_data %>% filter(n_agent %in% 1:5) %>% 
  ggplot() +
  aes(deltaBPM, theta, color = as.factor(n_agent)) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  )  + facet_wrap(~n_agent) + 
  coord_cartesian(xlim=c(-100,100)) # obs akser


```

# Prior predictive checks - pre stan

```{r}
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, alpha) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = alpha)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
 # scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - alpha",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )


# BETA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, beta) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = beta)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 40)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - beta",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )



# LAMBDA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, lambda) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = lambda)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 0.1)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - lambda",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )
```


## Getting into stan

```{r}
# compile model
file = file.path(stan_model_dir, 'simple_model_hrd_hierarchical.stan')
model <- cmdstan_model(
  file)

# make data stan friendly i.e. in list form 
hier_stan_data <- list(
  N = nrow(hier_data),
  S = 60,
  subj = hier_data$n_agent,
  dBPM = hier_data$deltaBPM,
  choice = hier_data$choice,
  prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
  prior_mu_alpha_scale         =  15, #changed 6/7 from 7 to 15
  prior_sigma_alpha_rate        =  0.1,    # E[sigma_alpha] = 1/rate = 10 BPM
  prior_mu_b_log_loc           =  2.5,    # exp(2.5) ~ 12 BPM slope
  prior_mu_b_log_scale          =  0.4,
  prior_sigma_b_log_rate        =  2.5,    # E[sigma_b_log] = 1/2.5 = 0.4
  prior_mu_lambda_logit_loc     = -3,      # 0.5 * inv_logit(-3) ~ 0.024
  prior_mu_lambda_logit_scale   =  0.4,
  prior_sigma_lambda_logit_rate =  2.5,    # E[sigma_lambda_logit] = 0.4
  run_diagnostics = 0L # not needed now 
)

str(hier_stan_data)
```
3. Sample

```{r}
# check if model has been run already
run_model <- regenerate_simulations || !file.exists(model_file)

if (!run_model) {
  fit <- readRDS(model_file)
}

if (run_model) {
  start_time <- Sys.time()

  pf <- model$pathfinder(data = hier_stan_data, num_paths = 10, draws = 400, psis_resample=FALSE)

  fit <- model$sample(
    data = hier_stan_data,
    init = pf,
    seed = 123,
    chains = 4,
    parallel_chains = 4,
    #threads_per_chain = 4,
    iter_warmup = 500,
    iter_sampling = 1000,
    refresh = 500#,
   # show_messages= FALSE
  )

  end_time <- Sys.time()
  time_equal <- end_time - start_time
  print(time_equal)
  
  fit$save_object(file = model_file)
}
```

Check diagnostics

```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")

```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit$summary(variables = c("mu_alpha", "mu_b_log", "mu_lambda_logit"))
fit_summary %>% dplyr::select(variable, mean, rhat, ess_bulk, ess_tail)
```




```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# Drop variables that are all NA (run_diagnostics = 0 quantities)
draws_df_clean <- draws_df |>
  select(where(~ !all(is.na(.x))))

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_alpha", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```

```{r}
# 3. Visual Checks (Trace and Rank Plots) - beta

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_b_log") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_b_log", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_lambda_logit") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_lambda_logit", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```


```{r}
# then get generated quantities
fit_gq <- model$generate_quantities(
  fitted_params = fit,
  data = list(
    N = nrow(hier_data),
    S = 60,
    subj = hier_data$n_agent,
    dBPM = hier_data$deltaBPM,
    choice = hier_data$choice,
    prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
    prior_mu_alpha_scale         =  15,
    prior_sigma_alpha_rate        =  0.1,    # E[sigma_alpha] = 1/rate = 10 BPM
    prior_mu_b_log_loc           =  2.5,    # exp(2.5) ~ 12 BPM slope
    prior_mu_b_log_scale          =  0.4,
    prior_sigma_b_log_rate        =  2.5,    # E[sigma_b_log] = 1/2.5 = 0.4
    prior_mu_lambda_logit_loc     = -3,      # 0.5 * inv_logit(-3) ~ 0.024
    prior_mu_lambda_logit_scale   =  0.4,
    prior_sigma_lambda_logit_rate =  2.5,    # E[sigma_lambda_logit] = 0.4
    run_diagnostics = 1L
  )
)
```

#2. Prior predictive checks

Do the prior allow only plausible behavior?


```{r}
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block
draws_gq <- as_draws_df(fit_gq$draws()) # Ensure we have the draws as a dataframe
observed_choice <- sum(hier_data$choice) # Total number of 'right' choices in the actual observed data

ppc_prior_ggplot <- ggplot(draws_gq, aes(x = choice_prior_sum)) +
  # Draw the histogram of the model's prior predictions (light red)
  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  
  # Add the vertical line for the actual observed data (dark red)
  geom_vline(xintercept = observed_choice, color = "darkred", linewidth = 1.2) +
  
  # Labels and theming
  ggtitle("Prior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = histogram spans the full range; observed value (dashed) is not in the tails") +
  xlab(paste("Total Number of 'Faster' Choices (out of", nrow(hier_data), "trials)")) + 
  ylab("Frequency") #+ 
  coord_cartesian(xlim = c(0,100))

ppc_prior_ggplot
```
ser ikke godt ud ! 

```{r}
draws_gq <- as_draws_df(fit_gq$draws())

# Extract alpha_prior and beta_prior for all subjects across draws
# and compute implied p(faster) at dBPM = 0 for each subject
draws_gq |>
  select(starts_with("alpha_prior"), starts_with("beta_prior"), starts_with("lambda_prior")) |>
  pivot_longer(everything(), names_to = c(".value", "s"),
               names_pattern = "(alpha_prior|beta_prior|lambda_prior)\\[(\\d+)\\]") |>
  mutate(p_faster = lambda_prior + (1 - 2 * lambda_prior) * pnorm((0 - alpha_prior) / beta_prior)) |>
  ggplot(aes(x = p_faster)) +
  geom_histogram(bins = 50, fill = "#ffbaba", colour = "#ff5252") +
  xlab("Prior p(faster) at dBPM = 0") +
  ggtitle("Prior predictive: per-subject p(faster)",
          subtitle = "Should be roughly uniform across [0,1] for a weakly informative prior")
```

```{r}
draws_gq |>
  select(starts_with("alpha_prior")) |>
  pivot_longer(everything(), names_to = "subject", values_to = "alpha_prior") |>
  ggplot(aes(x = alpha_prior)) +
  geom_histogram(bins = 80, fill = "#ffbaba", colour = "#ff5252") +
  xlab("Prior alpha (BPM threshold)") +
  ggtitle("Prior predictive: per-subject threshold",
          subtitle = "Should cover roughly [-30, +10] BPM for HRD task")
```


```{r}
draws <- fit_gq$draws("theta_prior_p", format = "matrix")

# Option 1: histogram of ALL theta_prior_p values (collapsed across trials and draws)
hist(as.vector(draws), breaks = 50, 
     main = "Prior predictive theta", xlab = "theta", xlim = c(0,1))

# Option 2: plot the mean theta per trial with uncertainty
theta_mean <- apply(draws, 2, mean)
theta_lower <- apply(draws, 2, quantile, 0.05)
theta_upper <- apply(draws, 2, quantile, 0.95)

plot(hier_data$deltaBPM, theta_mean, ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive psychometric function")
segments(hier_data$deltaBPM, theta_lower, hier_data$deltaBPM, theta_upper, col = "grey")

# Option 3: plot a random sample of draws as psychometric curves (most informative)
n_curves <- 50
sampled <- sample(nrow(draws), n_curves)

plot(NULL, xlim = range(hier_data$deltaBPM), ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive curves")
for (i in sampled) {
  lines(sort(hier_data$deltaBPM), draws[i, order(hier_data$deltaBPM)], col = rgb(0,0,1,0.1))
}
```

