---
title: "hierarchical model validation"
author: "Sigrid Bom"
date: "2026-07-06"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
# 1. Ensure 'pacman' is installed
if (!require("pacman")) install.packages("pacman")

# 2. Load required packages using pacman for efficient package management - ensures packages are installed if missing, then loads them
pacman::p_load(
  tidyverse,    # Core suite for data manipulation (dplyr) and plotting (ggplot2)
  here,         # For creating robust file paths relative to the project root (optional but good practice)
  posterior,    # Provides tools for working with posterior distributions
  cmdstanr,     # The R interface to Stan (via CmdStan)
  bayesplot,    # Specialized plotting functions for Bayesian models
  patchwork,    # For easily combining multiple ggplot plots
  priorsense,   # For automated prior sensitivity analysis
  SBC,          # Simulation-Based Calibration Checks
  future,       # Parallel processing
  furrr,        # Parallel mapping
  pracma,       # to be used in the probit regression formula
  cli,
  devtools,     # to download package from github
  progressr,     # to get progress bar in SBC
  remotes,       # to install specific version of ggplot needed to make priorsense dens plot
  msm
)

#devtools::install_github("hyunjimoon/SBC") # to make SBC
#progressr::handlers(global = TRUE) 
#progressr::handlers(c("rstudio", "progress")) # måske bedre??


# priorsense requires specific version of ggplot
#packageVersion("ggplot2")
#install.packages("remotes")
#remotes::install_version("ggplot2", version = "3.5.1")


# Set a default theme for ggplot for consistency
theme_set(theme_classic()) # Use a classic theme with white background

# Define relative paths 
stan_model_dir <- here("stan")
stan_results_dir <- here("simmodels")
sim_data_dir <- here("simdata")

# Create directories if they don't exist (important for first run)
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir)
if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)
if (!dir.exists(sim_data_dir)) dir.create(sim_data_dir)

# Define path for saving/loading the fitted model object that includes the generated quantities
#model_file <- file.path(stan_results_dir, "2_simple_hrd_model_hier_half_normal_priors.rds")
sbc_path <- file.path(stan_results_dir, "2_simple_hrd_model_hier_sbc_half_normal_priors.rds")

# Set to TRUE to rerun everything (takes time!), FALSE to load saved results.
regenerate_simulations <- FALSE 
#simulate_more_SBC = TRUE

# define functions

# probit regression - psychometric function
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
 }

# inverse logit
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```


Clean up of code in "hierarchical_hrd_model_sim_and_test.Rmd"

# Mathematical formulation

Each agent j {1, ..., J} has an unobserved parameter \theta_j on each trial i, which decides the choice _h_ of agent j on each trial i.

The theta parameter is found through a probit regression, which depends on several parameters including the stimulus level (delta BPM, x), the threshold (alpha), the slope (beta) and the lapse rate (lambda). All the parameters which are used to calculate theta are population level parameters, which means that they are drawn from a distrubition with a mean (mu) and standard deviation (sigma).

$$
h_i,_j \sim Bernoulli (\theta_i) \\
 \theta_i = \lambda + (1-2*\lambda)*(0.5+0.5*erf(\frac{x-\alpha}{\beta*\sqrt{2}}))
$$
The latent parameters are thought to be distributed as follows:
$$
\alpha_j \sim Normal(\mu_\alpha, \sigma_\alpha) \\
\beta log_j \sim Normal(\mu_\beta, \sigma_\beta) \\
\lambda  logit_j \sim Normal (\mu_\lambda, \sigma_\lambda)
$$
Overview of hyperparameters: (XX Need to update - changed sigma alpha and sigma b_log prior to be half normal)
$$
\mu_\alpha \sim Normal(-10,7) \\
\sigma_\alpha \sim Exponential(10) \\
\mu_\beta log \sim Normal(2.5,0.4) \\
\sigma_\beta log \sim Exponential(0.4)\\
\mu_\lambda logit \sim Normal(-3,0.4) \\
\sigma_\lambda logit \sim Exponential(0.4) \\
$$

Because beta needs to be positive, we sample it in the logarithmic space and then transform it back after sampling by exponentiating it. 
Similarly, we want lambda to be a small positive number, so we take the inverse logit of lambda and multiply it by 0.5.

Transformed parameters:
$$
\beta_j = exp(\beta log_j) \\
\lambda_j = 0.5* \frac{1}{(1+exp(-\lambda logit_j))}
$$

# 1. Generating model - same as before but now the priors are arguments in the function

```{r}
simulate_hierarchical_hrd_data <- function(n_agent = NULL,
                                           n_trials = NULL,
                                           prior_mu_alpha_loc = NULL,
                                           prior_mu_alpha_scale = NULL,
                                           prior_sigma_alpha_loc = NULL,
                                           prior_sigma_alpha_scale = NULL,
                                           prior_mu_b_log_loc = NULL,
                                           prior_mu_b_loc_scale = NULL,
                                           prior_sigma_b_log_loc = NULL,
                                           prior_sigma_b_log_scale = NULL,
                                           prior_mu_lambda_logit_loc = NULL,
                                           prior_mu_lambda_logit_scale = NULL,
                                           prior_sigma_lambda_logit_rate = NULL,
                                           seed = NULL
                                           ){
 
  if (!is.null(seed)) set.seed(seed)  # this generates a new seed for every run of the function if the seed is not set. so the only way to get the same results everytime is to set a seed when running the function (like sim_data(..., seed = 1997))
  
  # 1. sample population hyperparameters (if not provided)
  mu_alpha <- rnorm(1, mean = prior_mu_alpha_loc, sd = prior_mu_alpha_scale) # skal outputte omkring -10  # ændret fra 7 til 15 - 

  sigma_alpha <- abs(rnorm(1, mean = prior_sigma_alpha_loc, sd = prior_sigma_alpha_scale))
  # skal outputte omkring 10

  mu_b_log <- rnorm(1, mean = prior_mu_b_log_loc, sd = prior_mu_b_loc_scale) # skal outputte omkring 2.5 
  sigma_b_log <- abs(rnorm(1, mean = prior_sigma_b_log_loc, sd = prior_sigma_b_log_scale)) # skal outputte omrking 0.4 
  
  mu_lambda_logit <- rnorm(1, mean = prior_mu_lambda_logit_loc, sd = prior_mu_lambda_logit_scale) # skal være omrking -3
  sigma_lambda_logit <- rexp(1, rate = prior_sigma_lambda_logit_rate) #skal outputte omkring 0.4
  
  # 1.5 sample individual parameter values using hyper parameter values from population level
  
  data_sim = map_dfr(1:n_agent, ~ {
    # 1. sample true parameters per agent
    alpha         <- rnorm(1, mu_alpha, sigma_alpha)
    b_log         <- rnorm(1, mu_b_log, sigma_b_log)
    beta          <- exp(b_log)
    lambda_logit <- rnorm(1, mu_lambda_logit, sigma_lambda_logit)
    lambda       <- 0.5 * inv_logit(lambda_logit)
  
    # 2. sample stim
    deltaBPM <- msm::rtnorm(n_trials, mean = alpha, sd = 2 * beta, lower = -50, upper = 50) # truncated normal # is it a bad thing to depend the stimuli on the sampled parameters? also should this be different for each participant like it is now?
  
    # 3. calculate theta and choice
    theta_list <- probit_regression(x = deltaBPM, alpha = alpha, beta = beta, lambda = lambda)
    choice = rbinom(n=n_trials, size = 1, prob = theta_list) 
  
    # 4. return a tibble for each agent
    tibble(
      n_agent   = .x,
      n_trials,
      alpha,
      mu_alpha,
      sigma_alpha,
      beta,
      mu_b_log,
      sigma_b_log,
      lambda,
      mu_lambda_logit,
      sigma_lambda_logit,
      deltaBPM,
      theta = theta_list,
      choice
    )
})
  
 return(data_sim) 
}

```


```{r}
hier_data = simulate_hierarchical_hrd_data(n_agent = 60,
                                           n_trials = 40,
                                           prior_mu_alpha_loc = -10,
                                           prior_mu_alpha_scale = 7,
                                           prior_sigma_alpha_loc = 10,
                                           prior_sigma_alpha_scale = 3,
                                           prior_mu_b_log_loc = 2.5,
                                           prior_mu_b_loc_scale = 0.2,
                                           prior_sigma_b_log_loc = 0.4,
                                           prior_sigma_b_log_scale = 0.1,
                                           prior_mu_lambda_logit_loc = -3,
                                           prior_mu_lambda_logit_scale = 0.4,
                                           prior_sigma_lambda_logit_rate = 2.5,
                                           seed = 1985)

hier_data
```

Justification of prior values:
- We are running the data collection with 60 participants and 40 trials of the HRD. 
- *priors of mu_alpha*: we know from empirical data that participants tend to underestimate their heartrate with a threshold (alpha) of around -10, i.e., why the mean of the prior_mu_alpha is -10. Naturally, this varies, and the SD of 7 allows for variation. 
- *priors of sigma_alpha*: we want between-subject variation to allow for positive thresholds, too, why this is set to 15 (quite broad).

with alpha ~N(mu_alpha, sigma_alpha) and mu_alpha = -10, the probability an agent has an alpha > 0 is:


```{r}
1-pnorm(0, mean = -10, sd = 15)
```

This allows for 25% of agents (i.e., 15 agents out of 60) to have a positive threshold.

- *priors of mu_b_log and sigma_b_log*: 
beta is exp(b_log) and b_log ~ N(mu_b_log, sigma_b_log) per agent, so we want b_log to output around 2.5 +- 0.8 which we get by setting mu_b_log~ N(2.5, 0.2) and sigma_b_log ~N(0.4,1), so beta = exp(b_log) ranges roughly 5.5 to 27, all giving genuine sigmoid shapes over a ±50 stimulus range instead of degenerate step/flat functions.


```{r}
# plotting simulated data
hier_data %>% ggplot() + 
  aes(x = deltaBPM) + 
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Stim",
    y = "Density"
  )

# plotting the probabilities of answering 'faster'
hier_data %>%
  ggplot(aes(x = theta))  +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 colour = "black",
                 fill = "grey80") +
  geom_density(linewidth = 1) +
  labs(
    x = "Probability of answering 'faster'",
    y = "Density"
  )


hier_data %>% 
  ggplot() +
  aes(deltaBPM, theta, color = n_agent) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  ) + coord_cartesian(xlim=c(-80,80)) # obs akser

```


```{r fig.width=10, fig.height=5}
# visualise simulated data (choices)

hier_data %>% filter(n_agent %in% 1:5) %>% 
  ggplot() +
  aes(deltaBPM, theta, color = as.factor(n_agent)) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  )  + facet_wrap(~n_agent) 


```

# 2. Prior predictive checks - pre stan

```{r}
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    prior_mu_alpha_loc = -10,
    prior_mu_alpha_scale = 7,
    prior_sigma_alpha_loc = 10,
    prior_sigma_alpha_scale = 3,
    prior_mu_b_log_loc = 2.5,
    prior_mu_b_loc_scale = 0.2,
    prior_sigma_b_log_loc = 0.4,
    prior_sigma_b_log_scale = 0.1,
    prior_mu_lambda_logit_loc = -3,
    prior_mu_lambda_logit_scale = 0.4,
    prior_sigma_lambda_logit_rate = 2.5,
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, alpha) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = alpha)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
 # scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - alpha",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )


# BETA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, 
    prior_mu_alpha_loc = -10,
    prior_mu_alpha_scale = 7,
    prior_sigma_alpha_loc = 10,
    prior_sigma_alpha_scale = 3,
    prior_mu_b_log_loc = 2.5,
    prior_mu_b_loc_scale = 0.2,
    prior_sigma_b_log_loc = 0.4,
    prior_sigma_b_log_scale = 0.1,
    prior_mu_lambda_logit_loc = -3,
    prior_mu_lambda_logit_scale = 0.4,
    prior_sigma_lambda_logit_rate = 2.5,
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, beta) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = beta)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 40)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - beta",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )



# LAMBDA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agent = 100, 
    n_trials = 1, 
    prior_mu_alpha_loc = -10,
    prior_mu_alpha_scale = 7,
    prior_sigma_alpha_loc = 10,
    prior_sigma_alpha_scale = 3,
    prior_mu_b_log_loc = 2.5,
    prior_mu_b_loc_scale = 0.2,
    prior_sigma_b_log_loc = 0.4,
    prior_sigma_b_log_scale = 0.1,
    prior_mu_lambda_logit_loc = -3,
    prior_mu_lambda_logit_scale = 0.4,
    prior_sigma_lambda_logit_rate = 2.5,
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, lambda) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = lambda)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 0.1)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - lambda",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )
```

## Getting into stan

```{r}
# compile model
file = file.path(stan_model_dir, 'simple_model_hrd_hierarchical.stan')
model <- cmdstan_model(
  file)

# make data stan friendly i.e. in list form 
hier_stan_data <- list(
  N = nrow(hier_data),
  S = 60,
  subj = hier_data$n_agent,
  dBPM = hier_data$deltaBPM,
  choice = hier_data$choice,
  prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
  prior_mu_alpha_scale         = 7, #changed 6/7 from 7 to 15
  prior_sigma_alpha_loc        = 10,
  prior_sigma_alpha_scale      = 3,
  prior_mu_b_log_loc           = 2.5,    # exp(2.5) ~ 12 BPM slope
  prior_mu_b_log_scale         = 0.2,
  prior_sigma_b_log_loc        = 0.4,
  prior_sigma_b_log_scale      = 0.1,
  prior_mu_lambda_logit_loc    = -3,      # 0.5 * inv_logit(-3) ~ 0.024
  prior_mu_lambda_logit_scale  = 0.4,
  prior_sigma_lambda_logit_rate = 2.5,    # E[sigma_lambda_logit] = 0.4
  run_diagnostics = 0L # not needed now 
)


str(hier_stan_data)
```
# Sample

```{r}

n_iter_model = 3000
n_thin_model = 5
save_model = TRUE

# CHECK THIS 
model_file <- file.path(stan_results_dir, 
          sprintf("2_simple_hrd_model_hier_%diter_%dthin_half_normal_priors.rds", n_iter_model, n_thin_model))

# check if model has been run already
run_model <- regenerate_simulations || !file.exists(model_file)

if (!run_model) {
  fit <- readRDS(model_file)
}

if (run_model) {
  start_time <- Sys.time()

  pf <- model$pathfinder(data = hier_stan_data, num_paths = 10, draws = 400, psis_resample=FALSE)

  fit <- model$sample(
    data = hier_stan_data,
    init = pf,
    seed = 123,
    chains = 4,
    parallel_chains = 4,
    #threads_per_chain = 4,
    iter_warmup = 500,
    iter_sampling = n_iter_model,
    refresh = 500,
    thin = n_thin_model
   # show_messages= FALSE
  )

  end_time <- Sys.time()
  time_equal <- end_time - start_time
  print(time_equal)
  
  if (save_model){
    fit$save_object(file = model_file)
  }
}
```

## Check diagnostics

```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")

```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit$summary(variables = c("mu_alpha", "mu_b_log", "mu_lambda_logit", "sigma_alpha", "sigma_b_log", "sigma_lambda_logit"))
fit_summary %>% dplyr::select(variable, mean, rhat, ess_bulk, ess_tail)

draws_df <- as_draws_df(fit$draws())
```

Looks good. 


```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha

# Drop variables that are all NA (run_diagnostics = 0 quantities)
draws_df_clean <- draws_df |>
  select(where(~ !all(is.na(.x))))

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_alpha", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```

```{r}
# 3. Visual Checks (Trace and Rank Plots) - beta

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_b_log") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_b_log", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df_clean, pars = "mu_lambda_logit") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df_clean, pars = "mu_lambda_logit", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```

## Check acf
mcmc_acf(), mcmc_acf_bar()
Grid of autocorrelation plots by chain and parameter. The lags argument gives the maximum number of lags at which to calculate the autocorrelation function. mcmc_acf() is a line plot whereas mcmc_acf_bar() is a barplot.

```{r}
# A function to plot comparisons of thinning values See
# help("bayesplot_grid") for details on the bayesplot_grid function used here.
compare_fits_acf <- function(plot1, plot2, plot3, plot4, plot5, plot6, ncol = 2, ...) {
  bayesplot_grid(
    plot1, plot2, plot3, plot4, plot5, plot6,
    grid_args = list(
      ncol = ncol,
      top = sprintf("Autocorrelation functions: %d iter and chain thin = %d", n_iter_model, n_thin_model)
    ),
    #titles = "Autocorrelation function",
    #subtitles = sprintf("%d iter and chain thin = %d", n_iter_model, n_thin_model),
   # subtitles = c("Centered parameterization", 
    #              "Non-centered parameterization"),
    ...
  )# + labs(title="test")
}

```
```{r}

```


```{r}
# Drop variables that are all NA (run_diagnostics = 0 quantities)
draws_df <- as_draws_df(fit$draws())

draws_df_clean <- draws_df |>
  select(where(~ !all(is.na(.x))))

# ACF
 p_acf_mu_alpha <- bayesplot::mcmc_acf(draws_df_clean, pars = 'mu_alpha', lags = 10) #+ 
#   ggtitle("Autocorrelation function: mu_alpha", 
#           subtitle = sprintf("%d iter and chain thin = %d", n_iter_model, n_thin_model))

p_acf_sigma_alpha <- bayesplot::mcmc_acf(draws_df_clean, pars = 'sigma_alpha', lags = 10) #+ 
 
p_acf_mu_b_log <- bayesplot::mcmc_acf(draws_df_clean, pars = 'mu_b_log', lags = 10) #+ 

p_acf_sigma_b_log <- bayesplot::mcmc_acf(draws_df_clean, pars = 'sigma_b_log', lags = 10)

p_acf_mu_lambda_logit <- bayesplot::mcmc_acf(draws_df_clean, pars = 'mu_lambda_logit', lags = 10)# + 

p_acf_sigma_lambda_logit <- bayesplot::mcmc_acf(draws_df_clean, pars = 'sigma_lambda_logit', lags = 10)#

combined_acf_plot <- compare_fits_acf(p_acf_mu_alpha, p_acf_mu_b_log, p_acf_mu_lambda_logit, p_acf_sigma_alpha, p_acf_sigma_b_log, p_acf_sigma_lambda_logit, ncol=3)

combined_acf_plot
```
```{r}
save_acf_plots = FALSE
acf_plots_dir = file.path(here('plots', 'hierarchical_hrd_model', 'model_fits'))

if (save_acf_plots){
  ggsave(filename = sprintf("hier_hrd_acf_%diter_%dthin.png", 
        n_iter_model, n_thin_model), 
         plot = combined_acf_plot,
         path = acf_plots_dir,
         width = 16,
         height = 12,
         units = "cm",
         dpi = 300
         )
}
```


```{r}
neff <- neff_ratio(fit, pars = c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit"))

mcmc_neff(neff)
```


## Get generated quantities to prior predictive and posterior predictive

```{r}
# then get generated quantities
fit_gq <- model$generate_quantities(
  fitted_params = fit,
  data = list(
    N = nrow(hier_data),
    S = 60,
    subj = hier_data$n_agent,
    dBPM = hier_data$deltaBPM,
    choice = hier_data$choice,
    prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
    prior_mu_alpha_scale         = 7, #changed 6/7 from 7 to 15
    prior_sigma_alpha_loc        = 10,
    prior_sigma_alpha_scale      = 3,
    prior_mu_b_log_loc           = 2.5,    # exp(2.5) ~ 12 BPM slope
    prior_mu_b_log_scale         = 0.2,
    prior_sigma_b_log_loc        = 0.4,
    prior_sigma_b_log_scale      = 0.1,
    prior_mu_lambda_logit_loc    = -3,      # 0.5 * inv_logit(-3) ~ 0.024
    prior_mu_lambda_logit_scale  = 0.4,
    prior_sigma_lambda_logit_rate = 2.5,    # E[sigma_lambda_logit] = 0.4
    run_diagnostics = 1L # NOW we want diagnostics
  )
)
```

# 2. Prior predictive checks - using stan

Do the prior allow only plausible behavior?


```{r}
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block
draws_gq <- as_draws_df(fit_gq$draws()) # Ensure we have the draws as a dataframe
observed_choice <- sum(hier_data$choice) # Total number of 'right' choices in the actual observed data

ppc_prior_ggplot <- ggplot(draws_gq, aes(x = choice_prior_sum)) +
  # Draw the histogram of the model's prior predictions (light red)
  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  
  # Add the vertical line for the actual observed data (dark red)
  geom_vline(xintercept = observed_choice, color = "darkred", linewidth = 1.2) +
  
  # Labels and theming
  ggtitle("Prior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = histogram spans the full range; observed value (dashed) is not in the tails") +
  xlab(paste("Total Number of 'Faster' Choices (out of", nrow(hier_data), "trials)")) + 
  ylab("Frequency") +
  coord_cartesian(xlim = c(0,2400))

ppc_prior_ggplot
```
Central limit theorem - data will pile up around the center
observed choice sits in the bulk of the distribution
good: data spans almost entire possible range of data (from 0 to 2400)

```{r}
draws_gq <- as_draws_df(fit_gq$draws())

# Extract alpha_prior and beta_prior for all subjects across draws
# and compute implied p(faster) at dBPM = 0 for each subject
draws_gq |>
  select(starts_with("alpha_prior"), starts_with("beta_prior"), starts_with("lambda_prior")) |>
  pivot_longer(everything(), names_to = c(".value", "s"),
               names_pattern = "(alpha_prior|beta_prior|lambda_prior)\\[(\\d+)\\]") |>
  mutate(p_faster = lambda_prior + (1 - 2 * lambda_prior) * pnorm((0 - alpha_prior) / beta_prior)) |>
  ggplot(aes(x = p_faster)) +
  geom_histogram(bins = 50, fill = "#ffbaba", colour = "#ff5252") +
  xlab("Prior p(faster) at dBPM = 0") +
  ggtitle("Prior predictive: per-subject p(faster)",
          subtitle = "Should be roughly uniform across [0,1] for a weakly informative prior")



dBPM_grid <- unique(hier_data$deltaBPM)  # the actual stimulus values used

draws_gq |>
  select(starts_with("alpha_prior"), starts_with("beta_prior"), starts_with("lambda_prior")) |>
  pivot_longer(everything(), names_to = c(".value", "s"),
               names_pattern = "(alpha_prior|beta_prior|lambda_prior)\\[(\\d+)\\]") |>
  slice_sample(n = 200) |>  # subset for speed
  crossing(dBPM = dBPM_grid) |>
  mutate(p_faster = lambda_prior + (1 - 2 * lambda_prior) *
           pnorm((dBPM - alpha_prior) / beta_prior)) |>
  ggplot(aes(x = dBPM, y = p_faster, group = s)) +
  geom_line(alpha = 0.05, colour = "#ff5252") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  xlab("dBPM") + ylab("Prior p(faster)") +
  ggtitle("Prior predictive psychometric curves",
          subtitle = "Each line = one prior subject; should span a wide range of slopes and thresholds")
```

```{r}
draws_gq |>
  select(starts_with("alpha_prior")) |>
  pivot_longer(everything(), names_to = "subject", values_to = "alpha_prior") |>
  ggplot(aes(x = alpha_prior)) +
  geom_histogram(bins = 80, fill = "#ffbaba", colour = "#ff5252") +
  xlab("Prior alpha (BPM threshold)") +
  ggtitle("Prior predictive: per-subject threshold",
          subtitle = "Should cover roughly [-30, +10] BPM for HRD task")
```


```{r}
draws <- fit_gq$draws("theta_prior_p", format = "matrix")

# Option 1: histogram of ALL theta_prior_p values (collapsed across trials and draws)
hist(as.vector(draws), breaks = 50, 
     main = "Prior predictive theta", xlab = "theta", xlim = c(0,1))

# Option 2: plot the mean theta per trial with uncertainty
theta_mean <- apply(draws, 2, mean)
theta_lower <- apply(draws, 2, quantile, 0.05)
theta_upper <- apply(draws, 2, quantile, 0.95)

plot(hier_data$deltaBPM, theta_mean, ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive psychometric function")
segments(hier_data$deltaBPM, theta_lower, hier_data$deltaBPM, theta_upper, col = "grey")

# Option 3: plot a random sample of draws as psychometric curves (most informative)
n_curves <- 50
sampled <- sample(nrow(draws), n_curves)

plot(NULL, xlim = range(hier_data$deltaBPM), ylim = c(0,1),
     xlab = "dBPM", ylab = "theta",
     main = "Prior predictive curves")
for (i in sampled) {
  lines(sort(hier_data$deltaBPM), draws[i, order(hier_data$deltaBPM)], col = rgb(0,0,1,0.1))
}
```

# 3. Posterior predictive checks

```{r}
n_trials = 40

# Extract posterior predictive simulations from the fitted object
# This matrix has dimensions: [number_of_draws, number_of_trials]
choice_pred <- fit_gq$draws("choice_pred", format = "matrix")


# Calculate the test statistic T(y) for the *actual* empirical data
observed_choice <- sum(hier_data$choice) 

# Calculate the test statistic T(y_rep) for *each* replicated dataset
# We compute the row-wise sum across the posterior draws
choice_pred_sum <- rowSums(choice_pred)

# Bundle into a tibble for strictly structured plotting
ppc_df <- tibble(
  Replicated_Sum = choice_pred_sum
)

# --- Graphical Posterior Predictive Check using ggplot2 ---
ppc_posterior_ggplot <- ggplot(ppc_df, aes(x = Replicated_Sum)) +
  geom_histogram(fill = "skyblue", color = "white", alpha = 0.7, bins = 30) +
  geom_vline(xintercept = observed_choice, color = "black", linetype = "dashed", linewidth = 1) +
  ggtitle("Posterior Predictive Check: Total 'Faster' Choices",
          subtitle = "Healthy = observed value (dashed) sits inside the bulk of replicated datasets") +
  xlab(paste("Total Number of 'Faster' Choices (out of", n_trials, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_choice, y = Inf, 
    label = "Sim data", 
    vjust = 2, hjust = if(observed_choice > n_trials/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) 
ppc_posterior_ggplot
```

```{r}
ppc_overlay_df <- tibble(
  Replicated_Sum = c(draws_gq$choice_prior_sum, choice_pred_sum),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(choice_pred_sum)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 2. Generate the overlay plot
ppc_overlay_ggplot <- ggplot(ppc_overlay_df, aes(x = Replicated_Sum, fill = Distribution)) +
  # Use position = "identity" and alpha to ensure both distributions are visible
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 30) +
  geom_vline(xintercept = observed_choice, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Prior vs. Posterior Predictive: Narrowing of Predictions After Fitting",
          subtitle = "Posterior (blue) should be much tighter than prior (red) and centred on observed data") +
  xlab(paste("Total Number of 'Right' Choices (out of", n_trials, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_choice, y = Inf, 
    label = "Empirical data", 
    vjust = 2, hjust = if(observed_choice > n_trials/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(ppc_overlay_ggplot)
```
```{r}
# --- Example of checking another statistic (e.g., max run length) ---
# Define a function to calculate the maximum run length in a vector
max_run_length <- function(x) {
  if (length(x) == 0 || all(is.na(x))) return(0) 
  rle_x <- rle(x) # Run Length Encoding calculates consecutive identical choices
  if (length(rle_x$lengths) == 0) return(0) 
  max(rle_x$lengths) 
}

# 2. Calculate the empirical test statistic T(y)
observed_max_run <- max_run_length(hier_data$choice)

# 3. Extract prior replicated matrices and compute T(y_rep) for the prior
# y_prior_rep is an S x N matrix generated from the prior distributions
y_prior_rep <- fit_gq$draws("choice_prior_pred", format = "matrix")
prior_max_runs <- apply(y_prior_rep, 1, max_run_length)

# 4. Extract posterior replicated matrices and compute T(y_rep) for the posterior
# y_post_rep is an S x N matrix generated from the posterior distributions
y_post_rep <- fit_gq$draws("choice_pred", format = "matrix")
post_max_runs <- apply(y_post_rep, 1, max_run_length)

# 5. Bundle into a strictly structured tibble for overlay plotting
ppc_run_df <- tibble(
  Replicated_Max_Run = c(prior_max_runs, post_max_runs),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(post_max_runs)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 6. Generate the overlay plot
p_ppc_max_run <- ggplot(ppc_run_df, aes(x = Replicated_Max_Run, fill = Distribution)) +
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 20) +
  geom_vline(xintercept = observed_max_run, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Posterior Predictive: Maximum Run Length of Consecutive Choices",
          subtitle = "Tests sequential structure — a model assuming i.i.d. trials should still pass this") +
  xlab("Test Statistic T(y_rep): Maximum Run Length of Consecutive Choices") +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_max_run, y = Inf, 
    label = "Empirical T(y)", 
    vjust = 2, hjust = -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(p_ppc_max_run)
```
# 4. Prior-posterior update check

```{r}
# 4a. Look at the Posterior Estimate - alpha
ggplot(draws_df) +
  geom_density(aes(mu_alpha, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(mu_alpha_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = hier_data$mu_alpha, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_alpha (threshold)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "alpha  (threshold)",
    y        = "Density",
    fill     = "Distribution"
  ) 
```
```{r}
# 4b. Look at the Posterior Estimate - beta - tror noget går galt her
ggplot(draws_df) +
  geom_density(aes(exp(mu_b_log), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(exp(mu_b_log_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = exp(hier_data$mu_b_log), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_beta (slope)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "beta (slope)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```

```{r}
# 4c. Look at the Posterior Estimate - lambda
ggplot(draws_df) +
  geom_density(aes(inv_logit(mu_lambda_logit), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(inv_logit(mu_lambda_logit_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = inv_logit(hier_data$mu_lambda_logit), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_lambda (lapserate)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "lambda (lapserate)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```
# 5. Sensitivity analysis
# first combine draws from fit without and with generated quantities and remove empty variables from first fit
```{r}
# combine draws
draws_priorsense <- posterior::bind_draws(
  fit$draws(variables = c("mu_alpha", "sigma_alpha",
                          "mu_b_log", "sigma_b_log",
                          "mu_lambda_logit", "sigma_lambda_logit")),
  fit_gq$draws(variables = c("lprior_mu_alpha", "lprior_sigma_alpha",
                              "lprior_mu_b_log", "lprior_sigma_b_log",
                              "lprior_mu_lambda_logit", "lprior_sigma_lambda_logit",
                              "log_lik"))
) |> posterior::as_draws_df()

# get specs
sensitivity_summary <- priorsense::powerscale_sensitivity(draws_priorsense,
    prior_selection = c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit"))
sensitivity_summary

# We extract the variables of interest
# param_metrics <- sensitivity_summary |>
#   dplyr::filter(variable %in% c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit"))

# cat("\n--- Epistemic Diagnosis ---\n")
# print(param_metrics)

# plot
p_priorsense_dens <- priorsense::powerscale_plot_dens(
  draws_priorsense,
  prior_selection = c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log",
                      "mu_lambda_logit", "sigma_lambda_logit"),
  variables = c("mu_alpha", "sigma_alpha", "mu_b_log", "sigma_b_log",
                "mu_lambda_logit", "sigma_lambda_logit")
)
p_priorsense_dens
```
```{r}
# to get around memory issues
# draws_priorsense <- fit_gq$draws(
#   variables = c("mu_alpha", "sigma_alpha", 
#                 "lprior_mu_alpha", "lprior_sigma_alpha",
#                 "mu_b_log", "sigma_b_log",
#                 "lprior_mu_b_log", "lprior_sigma_b_log",
#                 "mu_lambda_logit", "sigma_lambda_logit",
#                 "lprior_mu_lambda_logit", "lprior_sigma_lambda_logit",
#                 "log_lik")
# ) |> posterior::as_draws_df()
# 
# p_priorsense_dens <- priorsense::powerscale_plot_dens(
#   draws_priorsense,
#   prior_selection = c("mu_alpha", "sigma_alpha","mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit"),
#   variables = c("mu_alpha", "sigma_alpha","mu_b_log", "sigma_b_log", "mu_lambda_logit", "sigma_lambda_logit")
# )
# p_priorsense_dens
```

# 6. Parameter recovery and SBC

SBC is heavy duty stuff. The SBC has therefore been run at a virtual cluster.

<!-- ```{r} -->
<!-- # Parallel SBC across simulations (SBC auto-balances cores per fit) -->
<!-- future::plan(multisession, workers = future::availableCores()) -->
<!-- options(SBC.min_chunk_size = 1) -->
<!-- cat(sprintf("Parallel SBC: %d workers\n", future::nbrOfWorkers())) -->

<!-- # Define the strict SBC generator wrapper -->
<!-- generate_sbc_dataset <- function(n_agent = 60, n_trials = 40) { -->

<!--   # 1. Simulate the universe (draws hyperparams and agent params from priors) -->
<!--   sim_data <- simulate_hierarchical_hrd_data( -->
<!--     n_agent = n_agent, -->
<!--     n_trials = n_trials -->
<!--   ) -->

<!--   sim_data <- sim_data %>% -->
<!--     rename( -->
<!--       S = n_agent, -->
<!--       N = n_trials -->
<!--       ) -->

<!--   # 2. Extract the true generative hyperparameters -->
<!--   mu_alpha_true <- sim_data$mu_alpha[1] -->
<!--   sigma_alpha_true <- sim_data$sigma_alpha[1] -->
<!--   mu_b_log_true <- sim_data$mu_b_log[1] -->
<!--   sigma_b_log_true <- sim_data$sigma_b_log[1] -->
<!--   mu_lambda_logit_true <- sim_data$mu_lambda_logit[1] -->
<!--   sigma_lambda_logit_true <- sim_data$sigma_lambda_logit[1] -->
<!--   #deltaBPM <- sim_data$deltaBPM -->

<!--   # 3. Extract the true individual latent parameters (must preserve agent order) -->
<!--   theta_true <- sim_data |> -->
<!--     distinct(S, theta) |> -->
<!--     arrange(S) |> -->
<!--     pull(theta) -->

<!--   # 4. Prepare the strictly typed list for CmdStanR -->
<!--   stan_data <- list( -->
<!--       N = nrow(sim_data), -->
<!--       S = n_agent, -->
<!--       subj = sim_data$S, -->
<!--       dBPM = sim_data$deltaBPM, -->
<!--       choice = sim_data$choice, -->
<!--       prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean -->
<!--       prior_mu_alpha_scale         =  7, -->
<!--       prior_sigma_alpha_rate        =  0.1,    # E[sigma_alpha] = 1/rate = 10 BPM -->
<!--       prior_mu_b_log_loc           =  2.5,    # exp(2.5) ~ 12 BPM slope -->
<!--       prior_mu_b_log_scale          =  0.4, -->
<!--       prior_sigma_b_log_rate        =  2.5,    # E[sigma_b_log] = 1/2.5 = 0.4 -->
<!--       prior_mu_lambda_logit_loc     = -3,      # 0.5 * inv_logit(-3) ~ 0.024 -->
<!--       prior_mu_lambda_logit_scale   =  0.4, -->
<!--       prior_sigma_lambda_logit_rate =  2.5,    # E[sigma_lambda_logit] = 0.4 -->
<!--       run_diagnostics = 0L # not needed now -->
<!-- ) -->

<!-- #str(hier_stan_data) -->

<!--   #stan_data <- prep_stan_data(sim_data, sim_data$n_agent) -->

<!--   # 5. Return the exact structure required by the SBC package -->
<!--   list( -->
<!--     variables = list(mu_alpha = mu_alpha_true, -->
<!--                      sigma_alpha = sigma_alpha_true, -->
<!--                      mu_b_log = mu_b_log_true, -->
<!--                      sigma_b_log = sigma_b_log_true, -->
<!--                      mu_lambda_logit = mu_lambda_logit_true, -->
<!--                      sigma_lambda_logit = sigma_lambda_logit_true -->
<!--                      ), -->
<!--     generated = stan_data -->
<!--   ) -->
<!--   # list( -->
<!--   #   variables = list( -->
<!--   #     mu_theta = mu_true, -->
<!--   #     sigma_theta = sigma_true, -->
<!--   #     theta_logit = theta_true -->
<!--   #   ), -->
<!--   #   generated = stan_data -->
<!--   # ) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- file = file.path(stan_model_dir, 'simple_model_hrd_hierarchical.stan') -->
<!-- model <- cmdstan_model( -->
<!--   file) -->

<!-- backend <- SBC::SBC_backend_cmdstan_sample( -->
<!--   model, -->
<!--   chains = 4, -->
<!--   parallel_chains = 4, -->
<!--   iter_warmup = 1000, -->
<!--   iter_sampling = 2000, -->
<!--   refresh = 0 -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # Register the generator function -->
<!-- generator <- SBC::SBC_generator_function(generate_sbc_dataset, n_agent = 60, n_trials = 40) -->

<!-- # Execute simulation based calibration -->
<!-- sbc_datasets <- SBC::generate_datasets(generator, 2) -->



<!-- start_time <-  Sys.time() -->

<!-- sbc_results <- SBC::compute_SBC(sbc_datasets, backend, keep_fits = FALSE) -->

<!-- end_time <- Sys.time() -->
<!-- time_equal <- end_time - start_time -->
<!-- print(time_equal) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # laod file from genome folder - virtual cluster -->
<!-- sbc_obj <- readRDS(here("scripts_genome", "genome_sbc_results", "2607152013_hierarchical_hrd_model_2000_sims_60subj_40trials_194m_53s_half_normal_COMBINED.rds")) -->
<!-- ``` -->



<!-- ```{r} -->
<!-- #sbc_obj <-readRDS(file.path(stan_results_dir, "2607101500_hierarchical_hrd_model_20_sims_60subj_40trials_5m_47s.rds")) -->
<!-- sbc_results <- sbc_obj$results -->
<!-- sbc_results -->
<!-- # no divergent transitions -->
<!-- # rhat is ok -->
<!-- # ess_bulk and ss_tail are good too -->
<!-- sbc_results$backend_diagnostics -->

<!-- hist(sbc_results$stats$rhat) -->

<!-- sbc_results$stats -->
<!-- sbc_datasets = sbc_obj$datasets -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # check ESS -->
<!-- ess_check <- sbc_results$stats |> -->
<!--   distinct(sim_id, variable, ess_bulk, ess_tail) |> -->
<!--   group_by(sim_id) |> -->
<!--   summarise(min_ess_bulk = min(ess_bulk, na.rm=TRUE), -->
<!--             min_ess_tail = min(ess_tail, na.rm=TRUE), .groups="drop") |> -->
<!--   arrange(min_ess_tail) -->

<!-- print(head(ess_check, 5)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- low_ess_ids <- c(8, 42, 14, 10, 33) -->

<!-- true_params_for_low_ess <- as.data.frame(sbc_datasets$variables) |> -->
<!--   mutate(sim_id = row_number()) |> -->
<!--   filter(sim_id %in% low_ess_ids) -->

<!-- print(true_params_for_low_ess) -->
<!-- ``` -->

<!-- ```{r} -->

<!-- bad_idx <- which(sbc_results$backend_diagnostics$n_divergent > 0 | sbc_results$default_diagnostics$max_rhat> 1.01) -->
<!-- sbc_datasets$variables[bad_idx, ] -->

<!-- ``` -->

```{r SBC results}
# laod file from genome folder - virtual cluster
sbc_obj <- readRDS(here("scripts_genome", "genome_sbc_results", "2607201406_hierarchical_hrd_model_3000iter_4thin_2000sims_60subj_40trials_124m_19s_half_normal_COMBINED.rds"))

results <- sbc_obj$results

# check results
results

# combine results into one tible
results_default_diag <- as_tibble(results$default_diagnostics)
results_backend_diag <- as_tibble(results$backend_diagnostics)
results_combined_diag <- left_join(results_default_diag, results_backend_diag, by = "sim_id")
```

```{r}
hist(results_combined_diag$min_ess_tail)
hist(results_combined_diag$min_ess_bulk)

```


Before we plot the results, we remove the bad fits.

```{r remove bad fits function}
#' Remove bad fits from an SBC combined diagnostics tibble
#'
#' Flags sim_ids that fail any of the following, using the combined
#' diagnostics tibble, then removes those sim_ids directly from the actual
#' SBC_results object (not just a diagnostics data frame) - so the returned
#' object can be passed straight into plot_rank_hist() / plot_ecdf() /
#' plot_ecdf_diff(), which require a proper SBC_results object rather than a
#' plain tibble.
#' @param sbc_obj The original SBC_results object (output of compute_SBC() /
#'   bind_results(), e.g. what you called `results` earlier).
#' @param results_combined_diag Tibble with one row per sim_id, containing
#'   columns: sim_id, max_rhat, n_divergent, n_rejects, ess_tail_ratio,
#'   min_ess_bulk, min_ess_tail (adjust names below if yours differ - run
#'   colnames(results_combined_diag) first to check).
#' @param rhat_threshold Max acceptable rhat (default 1.01)
#' @param min_ess_to_rank (tail ESS / max rank < 0.5) (default 0.5)
#'
#' @return A cleaned SBC_results object with the bad sim_ids removed - ready
#'   to pass directly into the SBC plotting functions.
remove_bad_fits <- function(sbc_obj,
                            results_combined_diag,
                             rhat_threshold = 1.01,
                             min_ess_to_rank_threshold = 0.5,
                             e_bfmi_threshold = 0.2,
                             min_ess_bulk_threshold = 400,
                             min_ess_tail_threshold = 400) {

  # --- 1. Build the "bad fit" logical condition, one entry per sim_id ---
  is_bad <- (results_combined_diag$n_divergent != 0) |
            (results_combined_diag$n_rejects != 0) |
            (results_combined_diag$max_rhat > rhat_threshold) |
            (results_combined_diag$min_ess_to_rank < min_ess_to_rank_threshold)
            (results_combined_diag$min_bfmi <= e_bfmi_threshold) |
            (results_combined_diag$min_ess_tail < min_ess_tail_threshold) |
            (results_combined_diag$min_ess_bulk < min_ess_bulk_threshold)

  # --- 2. Count and identify the bad sim_ids using sum() and which() ---
  n_total <- nrow(results_combined_diag)
  n_bad   <- sum(is_bad, na.rm = TRUE)
  pct_bad <- round(100 * n_bad / n_total, 1)

  bad_sim_ids <- results_combined_diag$sim_id[which(is_bad)]
  good_sim_ids <- results_combined_diag$sim_id[which(!is_bad)]


  # --- 3. Print a diagnostics summary ---
  cat(sprintf("Removed %d of %d fits (%.1f%%) for failing diagnostics.\n",
              n_bad, n_total, pct_bad))

  if (n_bad > 0) {
    cat("Removed sim_id's:\n")
    cat(paste(bad_sim_ids, collapse = ", "), "\n")

    # Optional: break down *why* each was flagged, useful for diagnosing
    # whether one criterion dominates (e.g. mostly divergences vs mostly rhat)
    breakdown <- results_combined_diag %>%
      filter(sim_id %in% bad_sim_ids) %>%
      summarise(
        n_divergent    = sum(n_divergent != 0, na.rm = TRUE),
        n_rejects      = sum(n_rejects != 0, na.rm = TRUE),
        n_high_rhat    = sum(max_rhat > rhat_threshold, na.rm = TRUE),
        n_min_ess_to_rank = sum(min_ess_to_rank < min_ess_to_rank_threshold, na.rm = TRUE),
        n_low_bfmi     = sum(min_bfmi <= e_bfmi_threshold, na.rm = TRUE),
        n_low_ess_tail = sum(min_ess_tail < min_ess_tail_threshold, na.rm = TRUE),
        n_low_ess_bulk = sum(min_ess_bulk < min_ess_bulk_threshold, na.rm = TRUE)
      )
    cat("\nBreakdown of failure reasons (fits can fail more than one criterion):\n")
    print(breakdown)
  }

  # --- 4. Subset the ORIGINAL SBC_results object directly, using SBC's own
  # `[` method for SBC_results objects. This keeps $stats, $default_diagnostics,
  # $backend_diagnostics etc. all internally consistent with each other - far
  # safer than trying to manually filter each component separately, since the
  # plotting functions expect a properly-formed SBC_results object.
  sbc_obj_clean <- sbc_obj[good_sim_ids]
 
  # Sanity check: confirm the subset actually has the expected number of sims
  # left. If this doesn't match, the indexing assumption above may not hold
  # for your SBC package version - worth checking nrow(sbc_obj_clean$stats)
  # against n_total - n_bad manually if this warning ever fires.
  n_remaining_check <- length(unique(sbc_obj_clean$stats$sim_id))
  if (n_remaining_check != (n_total - n_bad)) {
    warning(sprintf(
      "Expected %d sims remaining after subsetting, but SBC_results object has %d. Double check that sim_id indexing matches between results_combined_diag and sbc_obj.",
      n_total - n_bad, n_remaining_check
    ))
  }
  
  # --- 5. Report new minimum bulk/tail ESS after removing bad fits ---
  clean_diag <- results_combined_diag %>%
    filter(!is_bad)
 
  new_min_ess_bulk <- min(clean_diag$min_ess_bulk, na.rm = TRUE)
  new_min_ess_tail <- min(clean_diag$min_ess_tail, na.rm = TRUE)
 
  cat(sprintf(
    "\nAfter removal - new minimum bulk ESS: %.1f | new minimum tail ESS: %.1f\n",
    new_min_ess_bulk, new_min_ess_tail
  ))
  
   # --- 6. Attach the full report as an attribute on the cleaned object.
  # Using an attribute (rather than adding a new list element) means it
  # won't interfere with plot_rank_hist()/plot_ecdf()/etc, which only look
  # at the object's expected components - but you can still retrieve the
  # report anytime with attr(results_clean, "removed_fits_report").
  removed_fits_report <- list(
    n_total = n_total,
    n_bad = n_bad,
    pct_bad = pct_bad,
    bad_sim_ids = bad_sim_ids,
    good_sim_ids = good_sim_ids,
    breakdown = if (n_bad > 0) breakdown else NULL,
    new_min_ess_bulk = new_min_ess_bulk,
    new_min_ess_tail = new_min_ess_tail,
    thresholds = list(
      rhat_threshold = rhat_threshold,
      min_ess_to_rank_threshold = min_ess_to_rank_threshold,
      e_bfmi_threshold = e_bfmi_threshold,
      min_ess_tail_threshold = min_ess_tail_threshold,
      min_ess_bulk_threshold = min_ess_bulk_threshold
    )
  )
 
  attr(sbc_obj_clean, "removed_fits_report") <- removed_fits_report
 
  sbc_obj_clean
}
# ---------------------------------------------------------------------------
# Usage:
#
# clean_diag <- remove_bad_fits(results_combined_diag)
#
# good_sim_ids <- clean_diag$sim_id
# results_clean <- results[good_sim_ids]   # subset the SBC_results object
#
# plot_rank_hist(results_clean)
# plot_ecdf(results_clean)
# plot_ecdf_diff(results_clean)
# ---------------------------------------------------------------------------
```

```{r run remove bad fits function}
results_clean = remove_bad_fits(results, results_combined_diag)
```

```{r plot settings}
#plot directories
sbc_plots_dir = here("plots", "hierarchical_hrd_model", "parameter_recovery_and_SBC")

# SAVE PLOTS?
save_plots = TRUE
```


```{r param rec and SBC plots}
# to be used when plotting
report <- attr(results_clean, "removed_fits_report")
n_iterations = sbc_obj$n_iter 
n_thin = sbc_obj$n_thin # might have to set manually - not all model obj have this

# parameter recovery plot
p1 <- SBC::plot_sim_estimated(results_clean) + 
  ggtitle(
  label = sprintf("Parameter recovery (%d iter, thin = %d)", n_iterations, n_thin),
  subtitle = sprintf("Hierarchical HRD model w. %d fits of which %.0f bad fits were removed", report$n_total, report$n_bad)
 ) + 
  labs(x = "simulated value")

if (save_plots){
  ggsave(filename = sprintf("hier_hrd_param_rec_%diter_%dthin_%dsims.png", 
                            n_iterations, n_thin, report$n_total), 
         plot = p1, 
         path = sbc_plots_dir,
         width = 16,
         height = 10,
         units = "cm",
         dpi = 300
         )
}

print(p1)

# rank histogram
p2 <- SBC::plot_rank_hist(results_clean) + 
  ggtitle(
  label = sprintf("Simulation-based calibration check (%d iter, thin = %d)", n_iterations, n_thin),
  subtitle = sprintf("Hierarchical HRD model w. %d fits of which %.0f bad fits were removed", report$n_total, report$n_bad))

if (save_plots){
  ggsave(filename = sprintf("hier_hrd_rank_hist_%diter_%dthin_%dsims.png", 
                            n_iterations, n_thin, report$n_total), 
         plot = p2, 
         path = sbc_plots_dir,
         width = 16,
         height = 10,
         units = "cm",
         dpi = 300
         )
}

print(p2)

# ecdf plot
# empirical cumulative distribution function compared to the theoretical cumulative distribution function

p3 <- SBC::plot_ecdf(results_clean) + 
  ggtitle(
  label = sprintf("Empirical cumulative distribution function (ECDF) (%d iter, thin = %d)", n_iterations, n_thin),
    
  subtitle = sprintf("Hierarchical HRD model w. %d fits of which %.0f bad fits were removed", report$n_total, report$n_bad) 
  ) + 
  theme(legend.position = "bottom")

if (save_plots){
  ggsave(filename = sprintf("hier_hrd_ecdf_%diter_%dthin_%dsims.png", 
                            n_iterations, n_thin, report$n_total), 
         plot = p3, 
         path = sbc_plots_dir,
         width = 16,
         height = 10,
         units = "cm",
         dpi = 300
         )
}

print(p3)
  
# ecdf difference plot
# to more easily see if the empirical (simulated) CDF is well within the CDF 95% interval 

p4 <- SBC::plot_ecdf_diff(results_clean) + 
  ggtitle(
  label = sprintf("Difference between sampled and theoretical CDF (%d iter, thin = %d)", n_iterations, n_thin),
  subtitle = sprintf("Hierarchical HRD model w. %d fits of which %.0f bad fits were removed", report$n_total, report$n_bad)
  ) + 
    theme(
    legend.position = "bottom"
  )

if (save_plots){
  ggsave(filename = sprintf("hier_hrd_ecdf_diff_%diter_%dthin_%dsims.png", 
                            n_iterations, n_thin, report$n_total), 
         plot = p4, 
         path = sbc_plots_dir,
         width = 16,
         height = 12,
         units = "cm",
         dpi = 300
         )
}

print(p4)

report$breakdown
```

Parameter recovery of mu_alpha, sigma_alpha and mu_b_log look all right, mu_lambda_logit and sigma_lambda logit are more challenged, especially the sigma_lambda_logit. 

Simulation based calibration checks are ok-ish, but we have to remove about 17% of the model fits because of bad diagnostics. Especially the ESS values are challenged which indicates that the chains are not exploring the parameter space efficiently. 

The rank histograms look okay at first glance (equally distributed ranks) - except for the mu_lambda_logit parameter, where the model seems to overestimating the parameter values. However, when looking at the plots showing the difference between the sampled and the theoretical cumulative distriution function (CDF), it is evident that the model tends to underestimate the mu_b_log parameter and overestimate the sigma_b_log and mu_lambda_logit parameter (to a much greater extent in the mu_lambda_logit), which matches what we saw in the rank histogram plot. 

OBS! the min_Ess_to_rank check was added after the 3000 iter thin 4 was run. So all the other plots does not take this diagnostic into account when plotting the rank and parameter recovery results.