---
title: "data_generating_model"
author: "Sigrid Bom"
date: "2026-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,        # Suppress warnings
  message = FALSE,        # Suppress package loading messages
  echo = TRUE,           # Show R code
  fig.width = 8,         # Set default figure width
  fig.height = 5,        # Set default figure height
  fig.align = 'center',  # Center figures
  out.width = "80%",     # Make figures 80% of text width
  dpi = 300             # Set high resolution for figures
  )
pacman::p_load(tidyverse, cmdstanr, tidyverse, posterior, bayesplot, tidybayes, rstan, brms,pracma)

```

## Simulating data based on experimental design of the HRD (note RRST is a bit different)

from the BRMS demo:
*Note: To keep this simulation simple we do not use an adaptive method to optimize stimulus placement based on the responses of the participant. Instead, we simulate responses to 30 trials for each intensity within the pre-selected range (from -40 to +40 ΔBPM in increments of 5). This leads to a rather large total number of trials but the information contained in the dataset is not necessarily much more than in a dataset 10 times smaller with adaptive stimulus placement.*

```{r}
seed = 1997
set.seed(seed)

# study design
n_sub_hrdt <- 60
n_trials_hrdt <- 40
x_seq_hrdt <- rep(seq(-40, 40, 5),n_trials_hrdt) # check this

# Group-level parameters
# mu_lambda_HRDT <- -4.2
# sd_lambda_HRDT <- 2

mu_beta_hrdt <- -2.3
sd_beta_hrdt <- 0.3

mu_alpha_hrdt <- -8.6
sd_alpha_hrdt <- 11.6

mu_alpha_dif_hrdt = 5
mu_beta_dif_hrdt = 0

sd_beta_dif_hrdt <- sd_beta_hrdt * 0.5
sd_alpha_dif_hrdt <- sd_alpha_hrdt * 0.5

# Simulate subject-level parameters
subject_base_hrdt <- 
  tibble(
    subj = 1:n_subj_hrdt,
    alpha_base = rnorm(n_subj_hrdt, mu_alpha_hrdt, sd_alpha_hrdt),
  #  lambda_base = rnorm(n_subj_hrdt, mu_lambda_hrdt, sd_lambda_hrdt),
    beta_base = rnorm(n_subj_hrdt, mu_beta_hrdt, sd_beta_hrdt),
    alpha_effect = rnorm(n_subj_hrdt, mu_alpha_dif_hrdt, sd_alpha_dif_hrdt),
   # lambda_effect = 0,
    beta_effect = rnorm(n_subj_hrdt, mu_beta_dif_hrdt, sd_beta_dif_hrdt)
  )

# Expand to two conditions
condition_data_hrdt <- 
  subject_base_hrdt %>%
  mutate(
#    lambda_condition1 = inv_logit(lambda_base) / 2,
 #   lambda_condition2 = inv_logit(lambda_base + lambda_effect) / 2,

    beta_condition1 = exp(beta_base),
    beta_condition2 = exp(beta_base + beta_effect),

    alpha_condition1 = alpha_base,
    alpha_condition2 = alpha_base + alpha_effect
  ) %>%
  pivot_longer(
    cols = starts_with(c("alpha_", "beta_")),#, "lambda_")),
    names_to = c(".value", "condition"),
    names_pattern = "(.*)_condition(\\d)"
  ) %>% 
  drop_na() %>% 
  mutate(condition = factor(condition, levels = c("1", "2"), labels = c("control", "treatment")))

# Simulate trials
sim_data_hrdt <- 
  condition_data_hrdt %>%
  group_by(subj, condition) %>%
  mutate(x = list(x_seq_hrdt)) %>%
  unnest(x) %>%
  mutate(
    theta = (1 - 2 * lambda) * (0.5 + 0.5 * pracma::erf(beta * (x - alpha) / sqrt(2))),
    #theta = lambda + (1 - 2 * lambda) * (0.5 + 0.5 * pracma::erf(beta * (x - alpha) / sqrt(2))),
    y = rbinom(n(), 1, theta)
  ) %>%
  group_by(subj,condition,x,theta) %>% 
  summarise(n=sum(!is.na(y)),y=sum(y))

```

We want to start with a simple model:

$$
\alpha (threshold), \beta (slope) = f(choice,confidence, group, subject)
$$



```{r}

```

