---
title: "hierarchical hrd model"
author: "Sigrid Bom"
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load packages}
pacman::p_load(
  tidyverse,
  here,
  posterior,
  cmdstanr,
  brms, tidybayes, future, purrr, furrr, prettydoc, patchwork, stringr, ggpubr,
  pracma, dplyr
  )
```


```{r}
# functions

# probit regression - psychometric function
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
 }

# inverse logit
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```

Each agent j {1, ..., J} has an unobserved parameter \theta_j on each trial i, which decides the choice _h_ of agent j on each trial i.

The theta parameter is found through a probit regression, which depends on several parameters including the stimulus level (delta BPM, x), the threshold (alpha), the slope (beta) and the lapse rate (lambda). All the parameters which are used to calculate theta are population level parameters, which means that they are drawn from a distrubition with a mean (mu) and standard deviation (sigma).

$$
h_i,_j \sim Bernoulli (\theta_i) \\
 \theta_i = \lambda + (1-2*\lambda)*(0.5+0.5*erf(\frac{x-\alpha}{\beta*\sqrt{2}}))
$$
The latent parameters are thought to be distributed as follows:
$$
\alpha_j \sim Normal(\mu_\alpha, \sigma_\alpha) \\
\beta log_j \sim Normal(\mu_\beta, \sigma_\beta) \\
\lambda  logit_j \sim Normal (\mu_\lambda, \sigma_\lambda)
$$
Overview of hyperparameters:
$$
\mu_\alpha \sim Normal(-10,7) \\
\sigma_\alpha \sim Exponential(10) \\

\mu_\beta log \sim Normal(2.5,0.4) \\
\sigma_\beta log \sim Exponential(0.4)\\

\mu_\lambda logit \sim Normal(-3,0.4) \\
\sigma_\lambda logit \sim Exponential(0.4) \\
$$

Because beta needs to be positive, we sample it in the logarithmic space and then transform it back after sampling by exponentiating it. 
Similarly, we want lambda to be a small positive number, so we take the inverse logit of lambda and multiply it by 0.5.

Transformed parameters:
$$
\beta_j = exp(\beta log_j) \\
\lambda_j = 0.5* \frac{1}{(1+exp(-\lambda logit_j))}
$$

# 1. Generating model

```{r}
simulate_hierarchical_hrd_data <- function(n_agents = 60,
                                           n_trials = 40,
                                           mu_alpha = NULL,
                                           sigma_alpha = NULL,
                                           mu_b_log = NULL,
                                           sigma_b_log = NULL,
                                           mu_lambda_logit = NULL,
                                           sigma_lambda_logit = NULL,
                                           seed = NULL
                                           ){
 
  if (!is.null(seed)) set.seed(seed)    
  
  # 1. sample population hyperparameters (if not provided)
  if (is.null(mu_alpha)) mu_alpha <- rnorm(1, mean = -10, sd = 7) # skal outputte omkring -10
  if (is.null(sigma_alpha)) sigma_alpha <- rexp(1, rate = 1/10) # skal outputte omkring 10
  
  if (is.null(mu_b_log)) mu_b_log <- rnorm(1, mean = 2.5, sd = 0.4) # skal outputte omkring 2.5
  if (is.null(sigma_b_log)) sigma_b_log <- rexp(1, rate = 1/0.4) # skal outputte omrking 0.4
  
  if (is.null(mu_lambda_logit)) mu_lambda_logit <- rnorm(1, mean = -3, sd = 0.4) # skal være omrking -3
  if (is.null(sigma_lambda_logit)) sigma_lambda_logit <- rexp(1, rate = 1/0.4) #skal outputte omkring 0.4
  
  # 1.5 sample individual parameter values using hyper parameter values from population level
  
  data_sim = map_dfr(1:n_agents, ~ {
    # 1. sample true parameters per agent
    alpha         <- rnorm(1, mu_alpha, sigma_alpha)
    b_log         <- rnorm(1, mu_b_log, sigma_b_log)
    beta          <- exp(b_log)
    lambda_logit <- rnorm(1, mu_lambda_logit, sigma_lambda_logit)
    lambda       <- 0.5 * inv_logit(lambda_logit)
  
    # 2. sample stim
    deltaBPM <- rnorm(n_trials, alpha, 2 * beta) # is it a bad thing to depend the stimuli on the sampled parameters? also should this be different for each participant like it is now?
  
    # 3. calculate theta and choice
    theta_list <- probit_regression(x = deltaBPM, alpha = alpha, beta = beta, lambda = lambda)
    choice = rbinom(n=n_trials, size = 1, prob = theta_list) 
  
    # 4. return a tibble for each agent
    tibble(
      n_agent   = .x,
      alpha,
      mu_alpha,
      sigma_alpha,
      beta,
      mu_b_log,
      sigma_b_log,
      lambda,
      mu_lambda_logit,
      sigma_lambda_logit,
      deltaBPM,
      theta = theta_list,
      choice
    )
})
  
 return(data_sim) 
}

```


```{r}
# need to set seed to get same response every time
hier_data = simulate_hierarchical_hrd_data(seed = 123) 

hier_data
```


```{r}
hier_data %>% 
  ggplot() +
  aes(x =  hier_data$alpha_true) + geom_histogram()

 hier_data %>% 
  ggplot() +
  aes(x =  hier_data$beta_true) + geom_histogram()

 hier_data %>% 
  ggplot() +
  aes(x =  hier_data$lambda_true) + geom_histogram()
```

2. Compile stan model and make data stan friendly

```{r}
file = file.path('simple_model_hrd_hierarchical.stan')
model <- cmdstan_model(
  file)
```
```{r}
# make data stan friendly i.e. in list form 
hier_stan_data <- list(
  N = nrow(hier_data),
  S = 60,
  subj = hier_data$n_agent,
  dBPM = hier_data$deltaBPM,
  choice = hier_data$choice
)

str(hier_stan_data)
```

3. Sample

```{r}

start_time <- Sys.time()

fit <- model$sample(
  data = hier_stan_data,
  seed = 123,
  chains = 4,
  parallel_chains = 4,
  #threads_per_chain = 4,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 500#,
 # show_messages= FALSE
)

end_time <- Sys.time()
time_equal <- end_time - start_time
print(time_equal)
# Time difference of 
```

# Diagnostics check

```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")
```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit$summary(variables = c("alpha", "beta", "lambda"))
fit_summary %>% dplyr::select(variable, mean, rhat, ess_bulk, ess_tail)
```

```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "mu_alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "mu_alpha", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "mu_b_log") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "mu_b_log", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "mu_lambda_logit") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "mu_lambda_logit", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 4a. Look at the Posterior Estimate - alpha
ggplot(draws_df) +
  geom_density(aes(mu_alpha, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(mu_alpha_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = hier_data$mu_alpha, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_alpha (threshold)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "alpha  (threshold)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```
```{r}
# 4b. Look at the Posterior Estimate - beta - tror noget går galt her
ggplot(draws_df) +
  geom_density(aes(exp(mu_b_log), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(exp(mu_b_log_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = exp(hier_data$mu_b_log), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: Beta (slope)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "beta (slope)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```

```{r}
# 4c. Look at the Posterior Estimate - lambda
ggplot(draws_df) +
  geom_density(aes(inv_logit(mu_lambda_logit), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(inv_logit(mu_lambda_logit_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = inv_logit(hier_data$mu_lambda_logit), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: Lambda (lapserate)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "lambda (lapserate)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```

# OTHER notes below


```{r}
# purrr version
#theta_list <- map_dbl(deltaBPM, ~ probit_regression(x = .x, alpha = alpha_true, beta = beta_true, lambda=lambda_true))

# which is actually the same as
#theta_list2 <-probit_regression(x = deltaBPM, alpha = alpha_true, beta = beta_true, lambda=lambda_true)

```



