---
title: "hierarchical hrd model"
author: "Sigrid Bom"
date: "2026-04-13"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load packages}
pacman::p_load(
  tidyverse,
  here,
  posterior,
  cmdstanr,
  brms, tidybayes, future, purrr, furrr, prettydoc, patchwork, stringr, ggpubr,
  pracma, dplyr, msm
  )
```
```{r}
# want progress bars? 
pacman::p_load(progressr)
progressr::handlers(global = TRUE) # run this in the console
```

```{r}
# add to code
rerun_code <- FALSE

# if file exists or above is true, rerun code 
```


```{r}
# functions

# probit regression - psychometric function
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
 }

# inverse logit
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```

Each agent j {1, ..., J} has an unobserved parameter \theta_j on each trial i, which decides the choice _h_ of agent j on each trial i.

The theta parameter is found through a probit regression, which depends on several parameters including the stimulus level (delta BPM, x), the threshold (alpha), the slope (beta) and the lapse rate (lambda). All the parameters which are used to calculate theta are population level parameters, which means that they are drawn from a distrubition with a mean (mu) and standard deviation (sigma).

$$
h_i,_j \sim Bernoulli (\theta_i) \\
 \theta_i = \lambda + (1-2*\lambda)*(0.5+0.5*erf(\frac{x-\alpha}{\beta*\sqrt{2}}))
$$
The latent parameters are thought to be distributed as follows:
$$
\alpha_j \sim Normal(\mu_\alpha, \sigma_\alpha) \\
\beta log_j \sim Normal(\mu_\beta, \sigma_\beta) \\
\lambda  logit_j \sim Normal (\mu_\lambda, \sigma_\lambda)
$$
Overview of hyperparameters:
$$
\mu_\alpha \sim Normal(-10,7) \\
\sigma_\alpha \sim Exponential(10) \\

\mu_\beta log \sim Normal(2.5,0.4) \\
\sigma_\beta log \sim Exponential(0.4)\\

\mu_\lambda logit \sim Normal(-3,0.4) \\
\sigma_\lambda logit \sim Exponential(0.4) \\
$$

Because beta needs to be positive, we sample it in the logarithmic space and then transform it back after sampling by exponentiating it. 
Similarly, we want lambda to be a small positive number, so we take the inverse logit of lambda and multiply it by 0.5.

Transformed parameters:
$$
\beta_j = exp(\beta log_j) \\
\lambda_j = 0.5* \frac{1}{(1+exp(-\lambda logit_j))}
$$

# 1. Generating model

```{r}
simulate_hierarchical_hrd_data <- function(n_agents = 60,
                                           n_trials = 40,
                                           mu_alpha = NULL,
                                           sigma_alpha = NULL,
                                           mu_b_log = NULL,
                                           sigma_b_log = NULL,
                                           mu_lambda_logit = NULL,
                                           sigma_lambda_logit = NULL,
                                           seed = NULL
                                           ){
 
  if (!is.null(seed)) set.seed(seed)    
  
  # 1. sample population hyperparameters (if not provided)
  if (is.null(mu_alpha)) mu_alpha <- rnorm(1, mean = -10, sd = 7) # skal outputte omkring -10 
  if (is.null(sigma_alpha)) sigma_alpha <- rexp(1, rate = 1/10) # skal outputte omkring 10
  
  if (is.null(mu_b_log)) mu_b_log <- rnorm(1, mean = 2.5, sd = 0.4) # skal outputte omkring 2.5 
  if (is.null(sigma_b_log)) sigma_b_log <- rexp(1, rate = 1/0.4) # skal outputte omrking 0.4 
  
  if (is.null(mu_lambda_logit)) mu_lambda_logit <- rnorm(1, mean = -3, sd = 0.4) # skal være omrking -3
  if (is.null(sigma_lambda_logit)) sigma_lambda_logit <- rexp(1, rate = 1/0.4) #skal outputte omkring 0.4
  
  # 1.5 sample individual parameter values using hyper parameter values from population level
  
  data_sim = map_dfr(1:n_agents, ~ {
    # 1. sample true parameters per agent
    alpha         <- rnorm(1, mu_alpha, sigma_alpha)
    b_log         <- rnorm(1, mu_b_log, sigma_b_log)
    beta          <- exp(b_log)
    lambda_logit <- rnorm(1, mu_lambda_logit, sigma_lambda_logit)
    lambda       <- 0.5 * inv_logit(lambda_logit)
  
    # 2. sample stim
    deltaBPM <- rtnorm(n_trials, mean = alpha, sd = 2 * beta, lower = -50, upper = 50) # truncated normal # is it a bad thing to depend the stimuli on the sampled parameters? also should this be different for each participant like it is now?
  
    # 3. calculate theta and choice
    theta_list <- probit_regression(x = deltaBPM, alpha = alpha, beta = beta, lambda = lambda)
    choice = rbinom(n=n_trials, size = 1, prob = theta_list) 
  
    # 4. return a tibble for each agent
    tibble(
      n_agents   = .x,
      n_trials,
      alpha,
      mu_alpha,
      sigma_alpha,
      beta,
      mu_b_log,
      sigma_b_log,
      lambda,
      mu_lambda_logit,
      sigma_lambda_logit,
      deltaBPM,
      theta = theta_list,
      choice
    )
})
  
 return(data_sim) 
}

```


```{r}
# need to set seed to get same response every time
hier_data = simulate_hierarchical_hrd_data(seed = 123) 

hier_data
```

```{r}
hier_data %>% 
  ggplot() +
  aes(deltaBPM, theta, color = n_agent) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  ) + coord_cartesian(xlim=c(-80,80)) # obs akser

```


```{r fig.width=10, fig.height=5}
# visualise simulated data (choices)

hier_data %>% filter(n_agent %in% 1:5) %>% 
  ggplot() +
  aes(deltaBPM, theta, color = as.factor(n_agent)) +
  geom_jitter() +
  labs(
    x = "stimulus intensity",
    y = "probability of answering 'faster'",
    title = "HRD: simulated data for 60 agents with 40 trials"
  )  + facet_wrap(~n_agent) + 
  coord_cartesian(xlim=c(-100,100)) # obs akser


```

# Prior predictive checks

```{r}
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agents = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, alpha) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = alpha)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
 # scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - alpha",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )


# BETA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agents = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, beta) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = beta)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 40)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - beta",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )



# LAMBDA
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_hrd_data(
    n_agents = 100, 
    n_trials = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(n_agent, lambda) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = lambda)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 0.1)) + #, breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions - lambda",
    subtitle = "Healthy = populations differ? Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )
```


```{r}
hier_data %>% 
  ggplot() +
  aes(x =  alpha) + geom_histogram()

 hier_data %>% 
  ggplot() +
  aes(x =  beta) + geom_histogram()

 hier_data %>% 
  ggplot() +
  aes(x =  lambda) + geom_histogram()
```

2. Compile stan model and make data stan friendly

```{r}
file = file.path('simple_model_hrd_hierarchical.stan')
model <- cmdstan_model(
  file)
```

```{r}
# make data stan friendly i.e. in list form 
hier_stan_data <- list(
  N = nrow(hier_data),
  S = 60,
  subj = hier_data$n_agent,
  dBPM = hier_data$deltaBPM,
  choice = hier_data$choice,
  prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
  prior_mu_alpha_scale         =  7,
  prior_sigma_alpha_rate        =  0.1,    # E[sigma_alpha] = 1/rate = 10 BPM
  prior_mu_b_log_loc           =  2.5,    # exp(2.5) ~ 12 BPM slope
  prior_mu_b_log_scale          =  0.4,
  prior_sigma_b_log_rate        =  2.5,    # E[sigma_b_log] = 1/2.5 = 0.4
  prior_mu_lambda_logit_loc     = -3,      # 0.5 * inv_logit(-3) ~ 0.024
  prior_mu_lambda_logit_scale   =  0.4,
  prior_sigma_lambda_logit_rate =  2.5    # E[sigma_lambda_logit] = 0.4
)

str(hier_stan_data)
```

3. Sample

```{r}

start_time <- Sys.time()

pf <- model$pathfinder(data = hier_stan_data, num_paths = 10, draws = 400, psis_resample=FALSE)

fit <- model$sample(
  data = hier_stan_data,
  init = pf,
  seed = 123,
  chains = 4,
  parallel_chains = 4,
  #threads_per_chain = 4,
  iter_warmup = 500,
  iter_sampling = 1000,
  refresh = 500#,
 # show_messages= FALSE
)

end_time <- Sys.time()
time_equal <- end_time - start_time
print(time_equal)
# Time difference of 
```

# save results
```{r}
#saveRDS(fit, file = "../stan_results/hier_hrd_model_alpha_sd_7_beta_lambda_sd_0_4.rds") - doesn't save chains
#fit <- readRDS("../stan_results/hier_model_alpha_sd_10_beta_lambda_sd_0_5.rds")
# save object 

fit$save_object("../stan_results/hier_hrd_model_v2.RDS")
#test <- readRDS("../stan_results/hier_hrd_model_v2.RDS")
```


# Diagnostics check

```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")
```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit$summary(variables = c("alpha", "beta", "lambda"))
fit_summary %>% dplyr::select(variable, mean, rhat, ess_bulk, ess_tail)
```

```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "mu_alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "mu_alpha", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "mu_b_log") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "mu_b_log", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - alpha
draws_df <- as_draws_df(fit$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- bayesplot::mcmc_trace(draws_df, pars = "mu_lambda_logit") + 
  theme_classic() + 
  ggtitle("Trace Plot", 
          subtitle = "Healthy = all chains overlap and fluctuate around the same stable band")

# A healthy rank histogram has uniform bars across all chains
p_rank <- bayesplot::mcmc_rank_overlay(draws_df, pars = "mu_lambda_logit", ref_line = TRUE) + 
  theme_classic() + 
  ggtitle("Trank plot", subtitle = "Healthy = chains interleave uniformly around the reference line — no chain dominates any rank band")

p_trace / p_rank
```
```{r}
# 4a. Look at the Posterior Estimate - alpha
ggplot(draws_df) +
  geom_density(aes(mu_alpha, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(mu_alpha_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = hier_data$mu_alpha, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_alpha (threshold)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "alpha  (threshold)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```
```{r}
# 4b. Look at the Posterior Estimate - beta - tror noget går galt her
ggplot(draws_df) +
  geom_density(aes(exp(mu_b_log), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(exp(mu_b_log_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = exp(hier_data$mu_b_log), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_beta (slope)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "beta (slope)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```

```{r}
# 4c. Look at the Posterior Estimate - lambda
ggplot(draws_df) +
  geom_density(aes(inv_logit(mu_lambda_logit), fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(inv_logit(mu_lambda_logit_prior), fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = inv_logit(hier_data$mu_lambda_logit), linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title    = "Prior-Posterior Update: mu_lambda (lapserate)",
    subtitle = "Healthy = posterior (blue) is narrower than prior (red) and shifted toward the true value (dashed)",
    x        = "lambda (lapserate)",
    y        = "Density",
    fill     = "Distribution"
  ) +
  theme_classic()
```
Also need prior predictive checks. haven't made those. 

# parameter recovery

```{r}
# ALPHA
# 1. Extract the full summary (which defaults to including median, q5, and q95)
recovery_summary <- draws_df |> 
  posterior::subset_draws(variable = "alpha") |> 
  posterior::summarise_draws() |> 
  mutate(n_agent= as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")))

# 2. Extract the true generative parameters directly from the simulated dataset
true_agent_params <- hier_data |> 
  distinct(n_agent, n_trials, alpha)

# 3. Join the posterior estimates with the ground truth
recovery_df <- true_agent_params |> 
  left_join(recovery_summary, by = "n_agent")

# 4. Plot Recovery with Shrinkage Dynamics
ggplot(recovery_df, aes(x = alpha, y = median)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.5, color = "steelblue") +
  geom_point(aes(size = n_trials), color = "midnightblue", alpha = 0.8) +
  labs(
    title = "Parameter Recovery: Alpha",
    subtitle = "Healthy = points near diagonal (y = x)",
    x = "True Parameter Value",
    y = "Posterior Median Estimate",
    size = "Number of Trials (N_j)"
  )
```

```{r}
# BETA
# 1. Extract the full summary (which defaults to including median, q5, and q95)
recovery_summary <- draws_df |> 
  posterior::subset_draws(variable = "beta") |> 
  posterior::summarise_draws() |> 
  mutate(n_agent= as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")))

# 2. Extract the true generative parameters directly from the simulated dataset
true_agent_params <- hier_data |> 
  distinct(n_agent, n_trials, beta)

# 3. Join the posterior estimates with the ground truth
recovery_df <- true_agent_params |> 
  left_join(recovery_summary, by = "n_agent")

# 4. Plot Recovery with Shrinkage Dynamics
ggplot(recovery_df, aes(x = beta, y = median)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.5, color = "steelblue") +
  geom_point(aes(size = n_trials), color = "midnightblue", alpha = 0.8) +
  labs(
    title = "Parameter Recovery: Beta",
    subtitle = "Healthy = points near diagonal (y = x)",
    x = "True Parameter Value",
    y = "Posterior Median Estimate",
    size = "Number of Trials (N_j)"
  )
```
```{r}
# BETA
# 1. Extract the full summary (which defaults to including median, q5, and q95)
recovery_summary <- draws_df |> 
  posterior::subset_draws(variable = "lambda") |> 
  posterior::summarise_draws() |> 
  mutate(n_agent= as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")))

# 2. Extract the true generative parameters directly from the simulated dataset
true_agent_params <- hier_data |> 
  distinct(n_agent, n_trials, lambda)

# 3. Join the posterior estimates with the ground truth
recovery_df <- true_agent_params |> 
  left_join(recovery_summary, by = "n_agent")

# 4. Plot Recovery with Shrinkage Dynamics
ggplot(recovery_df, aes(x = lambda, y = median)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.5, color = "steelblue") +
  geom_point(aes(size = n_trials), color = "midnightblue", alpha = 0.8) +
  labs(
    title = "Parameter Recovery: Lambda",
    subtitle = "Healthy = points near diagonal (y = x)",
    x = "True Parameter Value",
    y = "Posterior Median Estimate",
    size = "Number of Trials (N_j)"
  )
```

```{r}
prep_stan_data <- function(df, agent_id) {
  df <- df |>
    arrange(id, t) |>
    mutate(
      agent_id   = as.integer(factor(id)),
      species_id = as.integer(factor(species))
    )
  species_of <- df |> distinct(agent_id, species_id) |>
    arrange(agent_id) |> pull(species_id)
  boundaries <- df |>
    mutate(row = row_number()) |>
    group_by(agent_id) |>
    summarize(start = min(row), stop = max(row), .groups = "drop") |>
    arrange(agent_id)
  list(
    N = nrow(df), J = max(df$agent_id), S = max(df$species_id),
    species_of = species_of, agent = df$agent_id,
    choice = df$choice, op_choice = df$op_choice,
    start = boundaries$start, stop = boundaries$stop,
    agent_lookup = df |> distinct(agent_id, id, species, species_id),
    species_lookup = df |> distinct(species_id, species) |> arrange(species_id)
  )
}
```


# SBC

```{r}
# Define the strict SBC generator wrapper
generate_sbc_dataset <- function(n_agents = 60, n_trials = 40) {
  
  # 1. Simulate the universe (draws hyperparams and agent params from priors)
  sim_data <- simulate_hierarchical_hrd_data(
    n_agents = n_agents, 
    n_trials = n_trials
  )
  
  sim_data <- sim_data %>% 
    rename(
      S = n_agents,
      N = n_trials
      )
  
  # 2. Extract the true generative hyperparameters
  mu_alpha_true <- sim_data$mu_alpha[1]
  sigma_alpha_true <- sim_data$sigma_alpha[1]
  mu_b_log_true <- sim_data$mu_b_log[1]
  sigma_b_log_true <- sim_data$sigma_b_log[1]
  mu_lambda_logit_true <- sim_data$mu_lambda_logit[1]
  sigma_lambda_logit_true <- sim_data$sigma_lambda_logit[1]
  #deltaBPM <- sim_data$deltaBPM
  
  # 3. Extract the true individual latent parameters (must preserve agent order)
  theta_true <- sim_data |> 
    distinct(S, theta) |> 
    arrange(S) |> 
    pull(theta)
  
  # 4. Prepare the strictly typed list for CmdStanR
  stan_data <- list(
      N = nrow(sim_data),
      S = 60,
      subj = sim_data$S,
      dBPM = sim_data$deltaBPM,
      choice = sim_data$choice,
      prior_mu_alpha_loc           = -10,     # HRD subjects systematically underestimate -> negative mean
      prior_mu_alpha_scale         =  7,
      prior_sigma_alpha_rate        =  0.1,    # E[sigma_alpha] = 1/rate = 10 BPM
      prior_mu_b_log_loc           =  2.5,    # exp(2.5) ~ 12 BPM slope
      prior_mu_b_log_scale          =  0.4,
      prior_sigma_b_log_rate        =  2.5,    # E[sigma_b_log] = 1/2.5 = 0.4
      prior_mu_lambda_logit_loc     = -3,      # 0.5 * inv_logit(-3) ~ 0.024
      prior_mu_lambda_logit_scale   =  0.4,
      prior_sigma_lambda_logit_rate =  2.5    # E[sigma_lambda_logit] = 0.4
)

#str(hier_stan_data)
  
  #stan_data <- prep_stan_data(sim_data, sim_data$n_agent)
  
  # 5. Return the exact structure required by the SBC package
  list(
    variables = list(mu_alpha = mu_alpha_true, 
                     sigma_alpha = sigma_alpha_true,
                     mu_b_log = mu_b_log_true, 
                     sigma_b_log = sigma_b_log_true,
                     mu_lambda_logit = mu_lambda_logit_true,
                     sigma_lambda_logit = sigma_lambda_logit_true
                     ),
    generated = stan_data
  )
  # list(
  #   variables = list(
  #     mu_theta = mu_true,
  #     sigma_theta = sigma_true,
  #     theta_logit = theta_true
  #   ),
  #   generated = stan_data
  # )
}
```

```{r}
file = file.path('simple_model_hrd_hierarchical.stan')
model <- cmdstan_model(
  file)

backend <- SBC::SBC_backend_cmdstan_sample(
  model,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 0
)
```

```{r}
# Register the generator function
generator <- SBC::SBC_generator_function(generate_sbc_dataset, n_agents = 60, n_trials = 40)

# Execute simulation based calibration
datasets <- SBC::generate_datasets(generator, 50)


start_time <-  Sys.time()

sbc_results <- SBC::compute_SBC(datasets, backend, keep_fits = FALSE)

end_time <- Sys.time()
time_equal <- end_time - start_time
print(time_equal)
```

```{r}
fit_check <- model$sample(data = generate_sbc_dataset()$generated, chains = 1, iter_sampling = 100)
fit_check$summary() |> pull(variable) |> head(10)
```


```{r}
sbc_results # to get an overview # same as summary(sbc_results)
sbc_results$errors
# no divergent transitions
# rhat is ok
# ess_bulk and ss_tail are good too
sbc_results$backend_diagnostics

hist(sbc_results$stats$rhat)

saveRDS(sbc_results, here('SBC', 'sbc_results_hier2.rds'))
```

```{r}
diagnostics <- sbc_results$default_diagnostics
diagnostics
sbc_results$stats
```
```{r}
SBC::plot_rank_hist(sbc_results)
```
```{r}
# parameter recovery 
SBC::plot_sim_estimated(sbc_results)
```

# OTHER notes below


```{r}
# purrr version
#theta_list <- map_dbl(deltaBPM, ~ probit_regression(x = .x, alpha = alpha_true, beta = beta_true, lambda=lambda_true))

# which is actually the same as
#theta_list2 <-probit_regression(x = deltaBPM, alpha = alpha_true, beta = beta_true, lambda=lambda_true)

```



