---
title: "parameter_recovery"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#first time installing cmdstanr?
#install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
install_cmdstan()
pacman::p_load(cli)
#pacman::p_load(remotes)
#remotes::install_version("cli", version = "3.6.5")
pacman::p_load(devtools)
devtools::install_github("hyunjimoon/SBC")
```

```{r}
# want progress bars? 
pacman::p_load(progressr)
progressr::handlers(global = TRUE) # run this in the console
```


```{r load packages}
pacman::p_load(
  tidyverse,
  here,
  posterior,
  cmdstanr,
  brms, tidybayes, future, purrr, furrr, prettydoc, patchwork, stringr, ggpubr,
  pracma
  )
```


```{r}
# functions

# probit regression - psychometric function - cumulative normal distribution to be precise
 probit_regression <- function(x, alpha, beta, lambda) {
  lambda + (1 - 2 * lambda) * (0.5+0.5*pracma::erf((x-alpha)/(beta*sqrt(2))))
 }

# inverse logit
inv_logit<-function(x){
  y=1/(1+exp(-x))
  return(y)
}
```

```{r}
# for saving stan results
model_file_1 <- here("..", "stan_results", "01_simpleHRD.rds")

```


# Parameter recovery - one participant

```{r}
simulate_data <- function(alpha, beta, lambda, N) {
  
  sim_data <- data.frame(
    deltaBPM = rnorm(N, 0, 10),
    subject = 1
  ) %>%
    mutate(trial = 1:n())
  
  sim_data$p_faster <- probit_regression(
    x = sim_data$deltaBPM,
    alpha = alpha,
    beta = beta,
    lambda = lambda
  )
  
  sim_data$choice <- rbinom(
    n = nrow(sim_data),
    size = 1,
    prob = sim_data$p_faster
  )
  
  list(
    N = N,
    dBPM = sim_data$deltaBPM,
    choice = sim_data$choice
  )
}
```

```{r}
trial_sizes <- c(50, 100)
n_rep <- 2

param_grid <- expand.grid(
  alpha  = c(-12, -8, -4),
  beta   = c(10, 15, 20),
  lambda = c(0.01, 0.05),
  N      = trial_sizes,
  rep    = 1:n_rep
)
```

```{r}
# specify stan file path and compile model 

file = file.path('simple_model_hrd.stan')
model <- cmdstan_model(
  file)#, 
#  cpp_options = list(stan_threads = FALSE), # enable threading - don't know what this means
#  stanc_options = list("O1")) # basic optimization - don't know how
```


```{r, message=FALSE, warning=FALSE, results='hide'}
results <- vector("list", nrow(param_grid))

for (i in seq_len(nrow(param_grid))) {
  
  cat("Running simulation", i, "of", nrow(param_grid), "\n")
  
  pars <- param_grid[i,]
  
  stan_data <- simulate_data(
    alpha  = pars$alpha,
    beta   = pars$beta,
    lambda = pars$lambda,
    N      = pars$N
  )
  
  fit <- model$sample(
    data = stan_data,
    seed = 123,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 4000,
    show_messages = FALSE,
    refresh = 0
  )
  
  #post <- fit$summary()
  draws_rec <- as_draws_df(fit$draws()) # same thing as running fit$summary() and getting the mean
  
  results[[i]] <- tibble(
    alpha_true  = pars$alpha,
    beta_true   = pars$beta,
    lambda_true = pars$lambda,
    N           = pars$N,
    rep         = pars$rep,
    alpha_est   = mean(draws_rec$alpha),
    beta_est    = mean(draws_rec$beta),
    lambda_est  = mean(draws_rec$lambda)
  )
}

recovery_results <- bind_rows(results)
```

# beta(1,1)

```{r}
# checking things
first_results = recovery_results

mean(draws_rec$alpha)
first_results
recovery_results
t=fit$summary()
t$mean[2]
fit$summary()
```

```{r}
recovery_results %>%
  group_by(N) %>%
  summarise(
    cor_alpha  = cor(alpha_true, alpha_est),
    cor_beta   = cor(beta_true, beta_est),
    cor_lambda = cor(lambda_true, lambda_est)
  )
```
```{r, message=FALSE, warning=FALSE, results='hide'}
ggplot(recovery_results, aes(alpha_true, alpha_est)) +
  geom_point(alpha = .5) +
  geom_smooth() +
  geom_line(aes(x=alpha_true, y = alpha_true), color = 'red') +
  facet_wrap(~N) + 
  labs(
    title = "Parameter recovery of alpha"
  )
```

```{r, message=FALSE, warning=FALSE, results='hide'}
recovery_results %>% ggplot() +
  aes(beta_true, beta_est) +
  geom_point(alpha = .5) +
  geom_smooth() +
  geom_line(aes(x=beta_true, y = beta_true), color = 'red') +
  facet_wrap(~N) + 
  labs(
    title = "Parameter recovery of beta"
  )
```
```{r, message=FALSE, warning=FALSE, results='hide'}
recovery_results %>% ggplot() +
  aes(lambda_true, lambda_est) +
  geom_point(alpha = .5) +
  geom_smooth() +
  geom_line(aes(x=lambda_true, y = lambda_true), color = 'red') +
  facet_wrap(~N) + 
  labs(
    title = "Parameter recovery of lambda"
  )
```


prøv ny beta prior inde på hans course notes - chapter 5 - gjorde ikke nogen nævneværdig forskel

```{r}
# Define the model
file = file.path('simple_model_hrd.stan')
model <- cmdstan_model(file)

# Define the backend engine
backend <- SBC::SBC_backend_cmdstan_sample(
  model,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 0
)

N = 100

# Define the generator
generate_one_dataset <- function(N){

  #parameters
  alpha_true = rnorm(n = 1, mean = -10, sd = 10) # kør bredere SD
  b_log_true = rnorm(n = 1, mean = 2.5, sd = 0.4)
  beta_true = exp(b_log_true)
  lambda_logit_true = rnorm(n = 1, mean = -3, sd = 0.4)
  lambda_true= 0.5*inv_logit(lambda_logit_true)
  
  # data 
  deltaBPM = rnorm(N, alpha_true, 2*beta_true)
  #estimerer psi - men uden at være psi algoritmen
  # psi = placement af stimuli - har ikke noget at gøre med PR og SBC
  # nice to have men overhovedet ikke need to have 
#  lambda_true = exp(lambda_log_true)
  
  # fprintf("alpha_true: %f\n", alpha_true)
  # fprintf("beta_true: %f\n", beta_true)
  # fprintf("lambda_true: %f\n", lambda_true)
  
  #probability of answering faster
  theta_prob_true = probit_regression(
    x=deltaBPM,
    alpha = alpha_true,
    beta = beta_true,
    lambda = lambda_true
  )
  #fprintf("theta_prob_true: %f\n", theta_prob_true)
  
  #simulate one dataset w N trials
  y_sim = rbinom(n=N, size = 1, prob = theta_prob_true) #why is n = 100 and size = 1? if n is number of observations and size is number of trials?
  list(
    variables = list(alpha = alpha_true, b_log = b_log_true, lambda_logit = lambda_logit_true),
    generated = list(N = N, dBPM = deltaBPM, choice = y_sim)
  )
}

# Register the generator function
generator <- SBC::SBC_generator_function(generate_one_dataset, N =100)

# Execute simulation based calibration
datasets <- SBC::generate_datasets(generator, n_sims = 1000)
sbc_results <- SBC::compute_SBC(datasets, backend, keep_fits = FALSE)
```
```{r}
# gem SBC object saveRDS
getwd()
saveRDS(sbc_results, here('SBC', 'sbc_results2.rds'))
cat("Model fit completed and saved to:",model_file_1, "\n")

#fit_bernoulli$save_object(file = model_file_bernoulli)

test = readRDS(here('SBC', 'sbc_results2.rds'))
```

# Check diagnostics 

```{r}
sbc_results # to get an overview # same as summary(sbc_results)

# no divergent transitions
# rhat is ok
# ess_bulk and ss_tail are good too
sbc_results$backend_diagnostics

hist(sbc_results$stats$rhat)
```

```{r}
diagnostics <- sbc_results$default_diagnostics
diagnostics
sbc_results$stats
```


```{r}
sum(sbc_results$default_diagnostics
sbc_results$backend_diagnostics
sum(sbc_results$backend_diagnostics$n_divergent)
sbc_results$warnings
```

```{r}
SBC::plot_rank_hist(sbc_results)
```
```{r}
# parameter recovery 
SBC::plot_sim_estimated(sbc_results)
```

## Parameter recovery group-level (simulating more than one participant)

```{r}
simulate_group_data <- function(alpha, beta, lambda, N, P) {
  
  map_dfr(1:P, function(s) {
    
    tibble(
      deltaBPM = rnorm(N, 0, 10),
      subject  = s,
      trial    = 1:N
    ) %>%
      mutate(
        p_faster = probit_regression(
          x      = deltaBPM,
          alpha  = alpha,
          beta   = beta,
          lambda = lambda
        ),
        choice = rbinom(N, 1, p_faster)
      )
    
  }) %>%
    {list(
      N       = nrow(.),
      S       = P,
      subj_id = .$subject,
      dBPM    = .$deltaBPM,
      choice  = .$choice
    )}
}
```

```{r}

trial_sizes <- c(50, 100)
n_rep <- 2

param_grid <- expand.grid(
  alpha  = c(-12, -8, -4),
  beta   = c(10, 15, 20),
  lambda = c(0.01, 0.05),
  N      = trial_sizes,
  P = c(30, 60),
  rep    = 1:n_rep
)

```

```{r}
# specify stan file path and compile model 

file_group_level = file.path('simple_model_hrd_group_level.stan')
model <- cmdstan_model(
  file_group_level)#, 
#  cpp_options = list(stan_threads = FALSE), # enable threading - don't know what this means
#  stanc_options = list("O1")) # basic optimization - don't know how
```


```{r, message=FALSE, warning=FALSE, results='hide'}
results <- vector("list", nrow(param_grid))

for (i in seq_len(nrow(param_grid))) {
  
  cat("Running simulation", i, "of", nrow(param_grid), "\n")
  
  pars <- param_grid[i,]
  
  stan_data <- simulate_group_data(
    alpha  = pars$alpha,
    beta   = pars$beta,
    lambda = pars$lambda,
    N      = pars$N,
    P      = pars$P
  )
  
  fit <- model$sample(
    data = stan_data,
    seed = 123,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 4000,
    show_messages = FALSE,
    refresh = 0
  )
  
  #post <- fit$summary()
  draws_rec <- as_draws_df(fit$draws()) # same thing as running fit$summary() and getting the mean
  
  results[[i]] <- tibble(
    alpha_true  = pars$alpha,
    beta_true   = pars$beta,
    lambda_true = pars$lambda,
    N           = pars$N,
    P           = pars$P,
    rep         = pars$rep,
    alpha_est   = mean(draws_rec$alpha),
    beta_est    = mean(draws_rec$beta),
    lambda_est  = mean(draws_rec$lambda)
  )
}

recovery_results <- bind_rows(results)
```

SBC version

```{r}
# Define the model
file = file.path('simple_model_hrd_group_level.stan')
model <- cmdstan_model(file)

# Define the backend engine
backend <- SBC::SBC_backend_cmdstan_sample(
  model,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 0
)

# --- GROUP SETTINGS ---
S <- 50      # number of participants per group dataset
N <- 100     # trials per participant
# total trials per dataset = S * N

# Define the generator
generate_group_dataset <- function(S, N){

  #parameters shared across all participants
  alpha_true = rnorm(n = 1, mean = -10, sd = 10) # kør bredere SD
  b_log_true = rnorm(n = 1, mean = 2.5, sd = 0.4)
  beta_true = exp(b_log_true)
  lambda_logit_true = rnorm(n = 1, mean = -3, sd = 0.4)
  lambda_true= 0.5*inv_logit(lambda_logit_true)
  
  # storage
  dBPM_all <- numeric(S*N)
  choice_all <- integer(S*N)

  idx <- 1

  for (s in 1:S) {
    # stimuli per participant
    deltaBPM <- rnorm(N, alpha_true, 2*beta_true)

    # choice probability
    theta_prob <- probit_regression(
      x = deltaBPM,
      alpha = alpha_true,
      beta = beta_true,
      lambda = lambda_true
    )

    # simulate binary choices
    y_sim <- rbinom(N, size = 1, prob = theta_prob)

    # place in joint data vector
    dBPM_all[idx:(idx+N-1)] <- deltaBPM
    choice_all[idx:(idx+N-1)] <- y_sim
    idx <- idx + N
  }

  list(
    variables = list(
      alpha = alpha_true,
      b_log = b_log_true,
      lambda_logit = lambda_logit_true
    ),
    generated = list(
      N = S*N,
      dBPM = dBPM_all,
      choice = choice_all
    )
  )
}


# Register generator
generator <- SBC::SBC_generator_function(
  generate_group_dataset,
  S = S,
  N = N
)

# Run SBC
datasets <- SBC::generate_datasets(generator, n_sims = 100)
sbc_results <- SBC::compute_SBC(datasets, backend, keep_fits = FALSE)
```

```{r}
# gem SBC object saveRDS
# for saving stan results
model_file_2 <- here("..", "stan_results", "01_simpleHRD_group_level.rds")
saveRDS(sbc_results, here('SBC', 'sbc_results_group.rds'))
cat("Model fit completed and saved to:",model_file_2, "\n")

#fit_bernoulli$save_object(file = model_file_bernoulli)

#sbc_results = readRDS(here('SBC', 'sbc_results_group.rds'))
```

# Check diagnostics 

```{r}
sbc_results # to get an overview # same as summary(sbc_results)

# no divergent transitions
# rhat is ok
# ess_bulk and ss_tail are good too
sbc_results$backend_diagnostics

hist(sbc_results$stats$rhat)
```

```{r}
# parameter recovery 
SBC::plot_sim_estimated(sbc_results)
```

```{r}
# sbc
SBC::plot_rank_hist(sbc_results)
```

next steps:
- hierarkisk - group mean og between subject variance (ikke kig så meget på subject level)
- eller psyk + confidence 


# Hierarkisk model

```{r}
# Define the model
file = file.path('simple_model_hrd_hierarchical.stan')
model <- cmdstan_model(file)

# Define the backend engine
backend <- SBC::SBC_backend_cmdstan_sample(
  model,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 0
)

# --- GROUP SETTINGS ---
S <- 50      # number of participants per group dataset
N <- 100     # trials per participant
# total trials per dataset = S * N

generate_hier_group_dataset <- function(S, N){

  # population-level parameters
  mu_alpha <- rnorm(1, -10, 10)
  mu_b_log <- rnorm(1, 2.5, 0.4)
  mu_lambda_logit <- rnorm(1, -3, 0.4)

  sigma_alpha <- abs(rnorm(1, 3, 1))
  sigma_b_log <- abs(rnorm(1, 0.3, 0.1))
  sigma_lambda_logit <- abs(rnorm(1, 0.3, 0.1))

  # storage
  dBPM_all <- numeric(S * N)
  choice_all <- integer(S * N)
  subj_all <- integer(S * N)

  idx <- 1

  for (s in 1:S) {

    # subject-level parameters
    alpha <- rnorm(1, mu_alpha, sigma_alpha)
    b_log_subj <- rnorm(1, mu_b_log, sigma_b_log)
    beta <- exp(b_log_subj)

    lambda_logit_subj <- rnorm(1, mu_lambda_logit, sigma_lambda_logit)
    lambda <- 0.5 * inv_logit(lambda_logit_subj)

    # stimuli depend on subject-level parameters (as before)
    deltaBPM <- rnorm(N, alpha, 2 * beta)

    theta_prob <- probit_regression(
      x = deltaBPM,
      alpha = alpha,
      beta = beta,
      lambda = lambda
    )

    y_sim <- rbinom(N, 1, theta_prob)

    # store
    dBPM_all[idx:(idx+N-1)] <- deltaBPM
    choice_all[idx:(idx+N-1)] <- y_sim
    subj_all[idx:(idx+N-1)] <- s

    idx <- idx + N
  }

  list(
    variables = list(
      mu_alpha = mu_alpha,
      mu_b_log = mu_b_log,
      mu_lambda_logit = mu_lambda_logit,
      sigma_alpha = sigma_alpha,
      sigma_b_log = sigma_b_log,
      sigma_lambda_logit = sigma_lambda_logit
    ),
    generated = list(
      N = S * N,
      S = S,
      dBPM = dBPM_all,
      choice = choice_all,
      subj = subj_all
    )
  )
}

# Register generator
generator <- SBC::SBC_generator_function(
  generate_hier_group_dataset,
  S = S,
  N = N
)

# Run SBC
datasets <- SBC::generate_datasets(generator, n_sims = 10)
sbc_results <- SBC::compute_SBC(datasets, backend, keep_fits = FALSE)
```




```{r}
# Define the generator
generate_group_dataset <- function(S, N){

  #parameters shared across all participants
  alpha_true = rnorm(n = 1, mean = -10, sd = 10) # kør bredere SD
  b_log_true = rnorm(n = 1, mean = 2.5, sd = 0.4)
  beta_true = exp(b_log_true)
  lambda_logit_true = rnorm(n = 1, mean = -3, sd = 0.4)
  lambda_true= 0.5*inv_logit(lambda_logit_true)
  
  # storage
  dBPM_all <- numeric(S*N)
  choice_all <- integer(S*N)

  idx <- 1

  for (s in 1:S) {
    # stimuli per participant
    deltaBPM <- rnorm(N, alpha_true, 2*beta_true)

    # choice probability
    theta_prob <- probit_regression(
      x = deltaBPM,
      alpha = alpha_true,
      beta = beta_true,
      lambda = lambda_true
    )

    # simulate binary choices
    y_sim <- rbinom(N, size = 1, prob = theta_prob)

    # place in joint data vector
    dBPM_all[idx:(idx+N-1)] <- deltaBPM
    choice_all[idx:(idx+N-1)] <- y_sim
    idx <- idx + N
  }

  list(
    variables = list(
      alpha = alpha_true,
      b_log = b_log_true,
      lambda_logit = lambda_logit_true
    ),
    generated = list(
      N = S*N,
      dBPM = dBPM_all,
      choice = choice_all
    )
  )
}


# Register generator
generator <- SBC::SBC_generator_function(
  generate_group_dataset,
  S = S,
  N = N
)

# Run SBC
datasets <- SBC::generate_datasets(generator, n_sims = 100)
sbc_results <- SBC::compute_SBC(datasets, backend, keep_fits = FALSE)
```

```{r}
# gem SBC object saveRDS
# for saving stan results
model_file_2 <- here("..", "stan_results", "01_simpleHRD_group_level.rds")
saveRDS(sbc_results, here('SBC', 'sbc_results_group.rds'))
cat("Model fit completed and saved to:",model_file_2, "\n")

#fit_bernoulli$save_object(file = model_file_bernoulli)

#sbc_results = readRDS(here('SBC', 'sbc_results_group.rds'))
```

# Check diagnostics 

```{r}
sbc_results # to get an overview # same as summary(sbc_results)

# no divergent transitions
# rhat is ok
# ess_bulk and ss_tail are good too
sbc_results$backend_diagnostics

hist(sbc_results$stats$rhat)
```


